---
title: "Gene expression project"
subtitle: "Analysis script"
author: "Bernice Sepers & Ane Liv Berthelsen"

format:
  html: 
  #pdf: 
    papersize: "a4paper"
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: FALSE
    documentclass: article
    geometry:
      - inner=3cm
      - outer=3cm
      - top=2cm
      - bottom=3cm
      - headsep=22pt
      - headheight=11pt
      - footskip=33pt
      - ignorehead
      - ignorefoot
      - heightrounded
    fig-pos: 'h'
    fig-cap-location: bottom
    tbl-cap-location: top
editor: visual
date: 11/20/2023
mainfont: Times New Roman
---

```{r}
install.packages("here")
install.packages("pacman")
library(here)
here::here()
```

```{r}
#| label: packages
#| echo: FALSE
#here::here()
invisible(pacman::p_load(dplyr, tidyverse, cowplot, edgeR, DESeq2, RColorBrewer, ggplot2, reshape2, ggpubr))

#cowplot for toggling with plots
#edgeR  loads limma as a dependency, limma for analysing microassay and RNA seq data
#DESeq2 needed for rowCounts
#reshape2 needed for the first plot
#ggpubr need to bind plots together
```

# Data 

```{r}
#| label: load_data
#| echo: FALSE
#Load in the filtered data
#Mutate all needed predictors into factors
d_n<-readRDS("rnaseq_filtered_Liv.Rdata")  
d_n$samples$life_history_stage <- as.factor(d_n$samples$life_history_stage)
d_n$samples$colony <- as.factor(d_n$samples$colony)
d_n$samples$ID <- as.factor(d_n$samples$ID)
d_n$samples$year <- as.factor(d_n$samples$year)
d_n$samples$group <- as.factor(d_n$samples$group)
d_n$samples$pair <- as.factor(d_n$samples$pair)
#dim(d_n) 12639 rows
```

# PCA
```{r}
#Use PCA to identify outliers. F24_FWB_mum_start_2018B was already removed at an earlier state
cpm_mat <- cpm(d_n, log=TRUE)  #we need the log values
exprs_mat <- cpm_mat
exprs_t <- t(exprs_mat) #transpose
pca_result <- prcomp(exprs_t, center = TRUE, scale. = TRUE)

pca_var <- pca_result$sdev^2                  
pca_var_explained <- pca_var / sum(pca_var)   
round(pca_var_explained[1:2] * 100, 2)

pc_scores <- as.data.frame(pca_result$x[, 1:2])  #keep PC1 and PC2
pc_scores$SampleID <- rownames(pc_scores)
pc_scores <- cbind(pc_scores, d_n$samples[match(pc_scores$SampleID, rownames(d_n$samples)), ])

PCA_LHS <- ggplot(pc_scores, aes(x = PC1, y = PC2, color = life_history_stage)) +
  geom_point(size = 3) +
  theme_classic() +
  stat_ellipse(type = "norm", size = 1) +
  xlim(-400, 150) +
  ylim(-150, 150) +
  labs(title = "PCA of gene expression profiles", x = "PC1 (24.15%)", y = "PC2 (12.38%)", color = "LHS")


PCA_LHS <- ggplot(pc_scores, aes(x = PC1, y = PC2, color = life_history_stage)) +
  geom_point(size = 3) +
  theme_classic() +
  stat_ellipse(type = "norm", size = 1) +
  xlim(-400, 150) +
  ylim(-150, 150) +
  labs(title = "PCA of gene expression profiles", x = "PC1 (24.15%)", y = "PC2 (12.38%)", color = "LHS")

PCA_TP <- ggplot(pc_scores, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 3) +
  theme_classic() +
  stat_ellipse(type = "norm", size = 1) +
  xlim(-400, 150) +
  ylim(-150, 150) +
  labs(title = "PCA of gene expression profiles", x = "PC1 (24.15%)", y = "PC2 (12.38%)", color = "TP")

PCA_plots <- ggarrange(PCA_LHS, PCA_TP)

#The PCA identifies F6_FWB_mum_start_2018B as an outlier

#Remove F6_FWB_mum_start_2018B and redo the PCA plots
d_n <- d_n[, d_n$samples$Sample != "F6_FWB_mum_start_2018B"]

#Re do the PCA analysis
cpm_mat <- cpm(d_n, log=TRUE)  #we need the log values
exprs_mat <- cpm_mat
exprs_t <- t(exprs_mat) #transpose
pca_result <- prcomp(exprs_t, center = TRUE, scale. = TRUE)

pca_var <- pca_result$sdev^2                  
pca_var_explained <- pca_var / sum(pca_var)   
round(pca_var_explained[1:2] * 100, 2)

pc_scores <- as.data.frame(pca_result$x[, 1:2])  #keep PC1 and PC2
pc_scores$SampleID <- rownames(pc_scores)
pc_scores <- cbind(pc_scores, d_n$samples[match(pc_scores$SampleID, rownames(d_n$samples)), ])

write.csv(pc_scores, file = here::here("GenesGO", "pc_scores.csv"))

#Plot in accordance to LHS and TP
PCA_LHS <- ggplot(pc_scores, aes(x = PC1, y = PC2, color = life_history_stage)) +
  geom_point(size = 2) +
  theme_classic() +
  stat_ellipse(type = "norm", size = 0.5) +
  scale_color_manual(values = c("#728C59", "#E3B778")) +
  xlim(-200, 200) +
  ylim(-200, 200) +
  labs(title = "c) PCA: Life history stage", x = "PC1 (23.12%)", y = "PC2 (13.25%)", color = NULL)

PCA_TP <- ggplot(pc_scores, aes(x = PC1, y = PC2, color = group)) +
  geom_point(size = 2) +
  theme_classic() +
  stat_ellipse(type = "norm", size = 0.5) +
  scale_color_manual(values = c("#728C59", "#E3B778")) +
  xlim(-200, 200) +
  ylim(-200, 200) +
  labs(title = "a) PCA: Time point", x = "PC1 (23.12%)", y = "PC2 (13.25%)", color = NULL)

PCA_Y <- ggplot(pc_scores, aes(x = PC1, y = PC2, color = year)) +
  geom_point(size = 2) +
  theme_classic() +
  stat_ellipse(type = "norm", size = 0.5) +
  scale_color_manual(values = c("#728C59", "#E3B778")) +
  xlim(-200, 200) +
  ylim(-200, 200) +
  labs(title = "c) PCA: Season", x = "PC1 (23.12%)", y = "PC2 (13.25%)", color = NULL)

PCA_C <- ggplot(pc_scores, aes(x = PC1, y = PC2, color = colony)) +
  geom_point(size = 2) +
  theme_classic() +
  stat_ellipse(type = "norm", size = 0.5) +
  scale_color_manual(values = c("#728C59", "#E3B778")) +
  xlim(-200, 200) +
  ylim(-200, 200) +
  labs(title = "d) PCA: Colony", x = "PC1 (23.12%)", y = "PC2 (13.25%)", color = NULL)

PCA_plots <- ggarrange(PCA_TP, PCA_LHS, PCA_Y, PCA_C)

ggsave(here::here("Plots","PCA_plots.png"), plot = PCA_plots, width = 7.2, height = 5, dpi = 600)
```

#Visualize data

```{r}
#| label: visualize_data
#| echo: FALSE
counts_per_million <- cpm(d_n)
log_counts_per_million <- cpm(d_n, log = T)
#Calculate mean and median value
L <- mean(d_n$samples$lib.size) * 1e-6 #average library size in millions
M <- median(d_n$samples$lib.size) * 1e-6 
#c(L, M) #19.72712, 19.11033
lcpm.cutoff <- log2(10/M + 2/L) #-0.678856

df_long <- data.frame(log_counts_per_million)
df_long$Gene <- rownames(df_long)
df_melted <- melt(df_long, id.vars = "Gene", variable.name = "Sample", value.name = "LogCPM")

#costum color palette
palette_200 <- colorRampPalette(brewer.pal(12, "Paired"))(200)

density_plot <- ggplot(df_melted, aes(x = LogCPM, color = Sample)) +
  geom_density(size = 0.5) + 
  scale_color_manual(values = palette_200) + 
  geom_vline(xintercept = lcpm.cutoff, linetype = "dashed", color = "black") +  # Add cutoff line
  labs(title = "Filtered Data", x = "Log-cpm", y = "Density") +
  theme_classic() +
  theme(legend.position = "none") +
  ylim(0,0.20)

density_plot

# Save the density plot as a PNG file
ggsave(here::here("Plots","density_plot.png"), plot = density_plot, width = 12, height = 8, dpi = 600)

#Remove elements not used in the rest of the script
rm(counts_per_million, density_plot, df_long, df_melted, log_counts_per_million, L, M, lcpm.cutoff, palette_200)
```
# Differential gene expression analysis

```{r}
clrs <- c("Upregulated" = "#FDA172", "Downregulated" = "#9867B5", "'Not significant'" = "grey20")
lfc <- 1 #log fold change of 1 is pretty standard but explore further
#a doubling in the original scaling is equal to a log2 fold change of 1
#a quadrupling is equal to a log2 fold change of 2
lfc_half <- 0.5
adj.P.value <- 0.05
fig_theme <- theme(text = element_text(family = "sans", size = 10),
                   legend.title = element_blank(),
                   legend.text=element_text(size=9),
                   axis.text.x = element_text(size = 9),
                   axis.text.y = element_text(size = 9))
```


## Time point: Analysis by Bernice
An example analysis where Bernice assessed differential expression per gene between the two time points (start vs end).

```{r}
##-------------- ANALYSIS OF TIME POINT --------------##
#Tagged out to not run dupcor again (the saved outcome can be called from the dupcor folder)
#design <- model.matrix(~ -1 + group, data = d_n$samples) #setting up model contrasts is more straight forward in the absence of an intercept for timepoint
#contr.matrix <- makeContrasts(
#  end_vs_start = group2-group1,
#  levels = colnames(design)) #create matrix for contrasts
# # Use voom to remove variance dependency on mean, see https://bioconductor.org/packages/release/workflows/vignettes/RNAseq123/inst/doc/limmaWorkflow.html
# # voom converts raw counts to log-CPM values by automatically extracting library sizes and normalisation factors from d itself. 
# #  If filtering of lowly-expressed genes is insufficient, a drop in variance levels can be observed at the low end of the expression scale due to very small counts.
# #voom() converts the read counts to log2-cpm, with associated weights, ready for linear modelling
#vobj_tp <- voom(d_n, design, plot=TRUE)
#
# # Fit a random effect
#dupcor_tp <- duplicateCorrelation(vobj_tp, design, block = d_n$samples$ID)
#dupcor_tp$consensus #the estimated intra individual correlation (0.3115568)
# #the intra individual correlation will change the voom weights slightly
# #run voom again considering the duplicateCorrelation results in order to compute more accurate precision weights
#vobj = voom(d_n, design, plot=TRUE, 
#            block=d_n$samples$ID, correlation=dupcor_tp$consensus)
# #update the correlation for the new voom weights
#dupcor_tp <- duplicateCorrelation(vobj, design, block = d_n$samples$ID)
#dupcor_tp$consensus #the estimated intra individual correlation (0.3114605)
#
# #Fit the model
#analyzing repeated measures data using duplicateCorrelation.
#The model forces the magnitude of the random effect to be the same across all genes.
# Estimate linear mixed model with a single variance component
# Fit the model for each gene
# But this step uses only the genome-wide average for the random effect
#vfit_tp <- lmFit(vobj, design, block=d_n$samples$ID, correlation=dupcor_tp$consensus)
#vfit_tp <- contrasts.fit(vfit_tp, contrasts=contr.matrix)
#
#Save the model to avoid recomputing everything
#save vfit_tp
#saveRDS(vfit_tp, file = here::here("Models","vfit_tp.rds"))
#
#Load vfit_tp
vfit_tp <- readRDS(here::here("Models","vfit_tp.rds"))

# Fit Empirical Bayes for moderated t-statistics, empirical Bayes moderation is carried out by borrowing information across all the genes to obtain more precise estimates of gene-wise variability
efit <- eBayes(vfit_tp)
topTable(efit, n=20)
plotSA(efit, main="Final model: Mean-variance trend")
plotMD(efit)
abline(h=0,col="darkgrey")

#adjusted p-value cutoff that is set at 5% by default.
efit$F
head(efit$F.p.value) #pval of the moderated F-statistics (equivalent to one-way ANOVA), it combines all contrast, so in this case it is the same as the t-statistic
head(efit$p.value) #same
summary(decideTests(efit, adjust.method="fdr"))
#set log-fold-changes (log-FCs) to be above a minimum value. The treat method (McCarthy and Smyth 2009) used to calculate p-values from empirical Bayes moderated t-statistics with a minimum log-FC requirement. 
#no need to use the ebayes output for this
#lfc<-0.14 #threshold set by Bernice to differentiate between differentially expressed genes

tfit <- treat(vfit_tp, lfc=lfc) #lfc = 1, lfc_half = 0.5
topTreat(tfit)
nrow(tfit)
dt <- decideTests(tfit, adjust.method="fdr")
summary(dt) 

#Test for significance
Allgenes_tp <- as.data.frame(topTreat(tfit, n=nrow(tfit)))
Filteredgenes_tp <- subset(Allgenes_tp, abs(logFC) >= lfc & adj.P.Val < adj.P.value)
Upregulated_tp <-subset(Filteredgenes_tp, Filteredgenes_tp$logFC >= lfc)
Downregulated_tp <-subset(Filteredgenes_tp, Filteredgenes_tp$logFC <= -lfc)
nrow(subset(Filteredgenes_tp, Filteredgenes_tp$logFC > 0))
nrow(subset(Filteredgenes_tp, Filteredgenes_tp$logFC < 0))

# Create a list to save the up and down regulated genes and the df containing all genes evaluated for GO analysis
tp_list <- list(All_tp = Allgenes_tp, Upregulated_tp = Upregulated_tp, Downregulated_tp = Downregulated_tp)

# Loop through and save each data frame using `here()`
for (name in names(tp_list)) {
  write.csv(tp_list[[name]], file = here::here("GenesGO", paste0(name, ".csv")), row.names = TRUE)
}

#Significance
Allgenes_tp$Significance <- "'Not significant'"
Allgenes_tp$Significance[Allgenes_tp$adj.P.Val < adj.P.value & Allgenes_tp$logFC >= lfc] <- "Upregulated"
Allgenes_tp$Significance[Allgenes_tp$adj.P.Val < adj.P.value & Allgenes_tp$logFC <= -lfc] <- "Downregulated"

#Plot volcano plot
tp_plot <- ggplot(Allgenes_tp, aes(x = logFC, y = -log10(P.Value), color = Significance, size = Significance)) +
  geom_point() +
  scale_color_manual(values = clrs) +  
  scale_size_manual(values = c(0.5, 1, 1)) +
  labs(title = "(a) Time point", x = "Log2 fold change", y = "-Log10 P-value") +
  theme_classic() +
  ylim(0,50) +
  xlim(-4,4) +
  theme(
    legend.position = "top",
    plot.title.position = "plot")
```
### Heatmap
```{r}
#Create heatmap for differentially expressed genes between time points
DEG_tp <- rownames(Filteredgenes_tp)

#Extract the expression values for the differentially expressed genes
expression_tp <- vobj$E[DEG_tp, ]

#get labels
annotation <- data.frame(TP = as.factor(d_n$samples$group))
rownames(annotation) <- d_n$samples$Sample
annotation$TP <- factor(annotation$TP, levels = c("1", "2"))

#order the expression values
matrix_tp <- expression_tp[, order(annotation$TP)]
annotation_tp <- annotation[order(annotation$TP), , drop = FALSE]

install.packages("pheatmap")  # only if you haven’t yet
library(pheatmap)

heatmap_tp <- pheatmap(matrix_tp,
                       scale = "row",  # normalize rows (genes)
                       color = colorRampPalette(c("#9D00FF", "white", "#FFA500"))(100),
                       annotation_colors = ann_colors,
                       cluster_rows = TRUE,
                       treeheight_row = 0, 
                       cluster_cols = FALSE,
                       gaps_col = cumsum(table(annotation$TP)),
                       show_rownames = TRUE,
                       show_colnames = FALSE,
                       annotation_legend = FALSE,
                       fontsize_col = 7,
                       fontsize_row = 6, 
                       main = "Time point")

ggsave(here::here("Plots","heatmap_tp.png"), plot = heatmap_tp, width = 3, height = 7, dpi = 600)
```

## Life stages

```{r}
#Tagged out to not run dupcor again (the saved outcome can be called from the dupcor folder)
# #Create a design matrix
#design_lhs <- model.matrix(~ -1 + life_history_stage, data = d_n$samples) #model contrasts without intercept is easier
#colnames(design_lhs) <- gsub("life_history_stage", "", colnames(design_lhs)) #clean group names
#
# #create matrix for contrasts
#contr.matrix.lhs <- makeContrasts(
#  adultvspup = Mother-Pup,
#  levels = colnames(design_lhs)) 
#
# #voom() converts the read counts to log2-cpm, with associated weights, ready for linear modelling
#vobj_lhs <- voom(d_n, design_lhs, plot=TRUE)
# 
# # Fit a random effect
#dupcor_lhs <- duplicateCorrelation(vobj_lhs, design_lhs, block = d_n$samples$ID)
#dupcor_lhs$consensus #the estimated intra individual correlation (0.1362346)
#
#vobj_lhs = voom(d_n, design_lhs, plot=TRUE, 
#          block=d_n$samples$ID, correlation=dupcor_lhs$consensus)
# #update the correlation for the new voom weights
#dupcor_lhs <- duplicateCorrelation(vobj_lhs, design_lhs, block = d_n$samples$ID)
#dupcor_lhs$consensus #the estimated intra individual correlation (0.1362741)
# #Analyzing repeated measures data using duplicateCorrelation.
# #The model forces the magnitude of the random effect to be the same across all genes.
#
#
# #Estimate linear mixed model with a single variance component
#vfit_lhs <- lmFit(vobj_lhs, design_lhs, block=d_n$samples$ID, correlation=dupcor_lhs$consensus)
#vfit_lhs <- contrasts.fit(vfit_lhs, contrasts=contr.matrix.lhs)
#
# #Save the model to avoid recomputing everything
# #save vfit_lhs
#saveRDS(vfit_lhs, file = here::here("Models","vfit_lhs.rds"))
#
#Load vfit_lhs
vfit_lhs <- readRDS(here::here("Models","vfit_lhs.rds"))

# Fit Empirical Bayes for moderated t-statistics, empirical Bayes moderation is carried out by borrowing information across all the genes to obtain more precise estimates of gene-wise variability
efit_lhs <- eBayes(vfit_lhs)
topTable(efit_lhs, n=20)
plotSA(efit_lhs, main="Final model: Mean-variance trend")
plotMD(efit_lhs)
abline(h=0,col="darkgrey")

#adjusted p-value cutoff that is set at 5% by default.
efit_lhs$F
#head(efit_lhs$F.p.value) #pval of the moderated F-statistics (equivalent to one-way ANOVA), it combines all contrast, so in this case it is the same as the t-statistic
#head(efit_lhs$p.value) #same
summary(decideTests(efit_lhs, adjust.method="fdr"))
#set log-fold-changes (log-FCs) to be above a minimum value. The treat method (McCarthy and Smyth 2009) used to calculate p-values from empirical Bayes moderated t-statistics with a minimum log-FC requirement. 
#no need to use the ebayes output for this

#Positive values means the gene is more expressed
tfit_lhs <- treat(vfit_lhs, lfc=lfc) 
topTreat(tfit_lhs)
nrow(tfit_lhs)
dt_lhs <- decideTests(tfit_lhs, adjust.method="fdr")
summary(dt_lhs) 

#Test for significance
Allgenes_lhs <- as.data.frame(topTreat(tfit_lhs, n=nrow(tfit_lhs)))
Filteredgenes_lhs <- subset(Allgenes_lhs, abs(logFC) >= lfc & adj.P.Val < adj.P.value)
Upregulated_lhs <-subset(Filteredgenes_lhs, Filteredgenes_lhs$logFC >= lfc)
Downregulated_lhs <-subset(Filteredgenes_lhs, Filteredgenes_lhs$logFC <= -lfc)
nrow(subset(Filteredgenes_lhs, Filteredgenes_lhs$logFC > 0))
nrow(subset(Filteredgenes_lhs, Filteredgenes_lhs$logFC < 0))

# Create a list to save the up and down regulated genes and the df containing all genes evaluated for GO analysis
lhs_list <- list(All_lhs = Allgenes_lhs, Upregulated_lhs = Upregulated_lhs, Downregulated_lhs = Downregulated_lhs)

# Loop through and save each data frame using `here()`
for (name in names(lhs_list)) {
  write.csv(lhs_list[[name]], file = here::here("GenesGO", paste0(name, ".csv")), row.names = TRUE)
}


#Significance
Allgenes_lhs$Significance <- "'Not significant'"
Allgenes_lhs$Significance[Allgenes_lhs$adj.P.Val < adj.P.value & Allgenes_lhs$logFC >= lfc] <- "Upregulated"
Allgenes_lhs$Significance[Allgenes_lhs$adj.P.Val < adj.P.value & Allgenes_lhs$logFC <= -lfc] <- "Downregulated"

#Plot DEG volcano plot
lhs_plot <- ggplot(Allgenes_lhs, aes(x = logFC, y = -log10(P.Value), color = Significance, size = Significance)) +
  geom_point() +
  scale_color_manual(values = clrs) +  
  scale_size_manual(values = c(0.5, 1, 1)) +
  labs(title = "(b) Life history stage", x = "Log2 fold change", y = "-Log10 P-value") +
  theme_classic() +
  #ylim(0,60) +
  #xlim(-4,4) +
    theme(
    legend.position = "top",
    plot.title.position = "plot")
```
### Heatmap
```{r}
#Create heatmap for differentially expressed genes between time points
DEG_lhs <- rownames(Filteredgenes_lhs)

#Extract the expression values for the differentially expressed genes
expression_lhs <- vobj_lhs$E[DEG_lhs, ]

#get labels
annotation <- data.frame(LHS = as.factor(d_n$samples$life_history_stage))
rownames(annotation) <- d_n$samples$Sample
annotation$LHS <- factor(annotation$LHS, levels = c("Mother", "Pup"))

#order the expression values
matrix_lhs <- expression_lhs[, order(annotation$LHS)]
annotation_lhs <- annotation[order(annotation$LHS), , drop = FALSE]

heatmap_lhs <- pheatmap(matrix_lhs,
                       scale = "row",  # normalize rows (genes)
                       color = colorRampPalette(c("#9D00FF", "white", "#FFA500"))(100),
                       annotation_colors = ann_colors,
                       cluster_rows = TRUE,
                       treeheight_row = 0, 
                       cluster_cols = FALSE,
                       gaps_col = cumsum(table(annotation$LHS)),
                       show_rownames = TRUE,
                       show_colnames = F,
                       annotation_legend = FALSE,
                       legend = F, 
                       fontsize_col = 7,
                       fontsize_row = 6, 
                       main = "Life history stage")

ggsave(here::here("Plots","heatmap_lhs.png"), plot = heatmap_lhs, width = 7, height = 11, dpi = 600)
```

## Colony

```{r}
#Tagged out to not run dupcor again (the saved outcome can be called from the dupcor folder)
# #Create a design matrix
#design_c <- model.matrix(~ -1 + colony, data = d_n$samples) #model contrasts without intercept is easier
#colnames(design_c) <- gsub("colony", "", colnames(design_c)) #clean group names
#
# #create matrix for contrasts
#contr.matrix.c <- makeContrasts(
#  SSBvsFWB = SSB-FWB,
#  levels = colnames(design_c)) 
#
# #voom() converts the read counts to log2-cpm, with associated weights, ready for linear modelling
#vobj_c <- voom(d_n, design_c, plot=TRUE)
#
# # Fit a random effect
#dupcor_c <- duplicateCorrelation(vobj_c, design_c, block = d_n$samples$ID)
#dupcor_c$consensus #the estimated intra individual correlation (0.1998915)
# #the intra individual correlation will change the voom weights slightly
# #run voom again considering the duplicateCorrelation results in order to compute more accurate precision weights
#vobj_c = voom(d_n, design_c, plot=TRUE, 
#            block=d_n$samples$ID, correlation=dupcor_c$consensus)
# #update the correlation for the new voom weights
#dupcor_c <- duplicateCorrelation(vobj_c, design_c, block = d_n$samples$ID)
#dupcor_c$consensus #the estimated intra individual correlation (0.1999114)
#
# #Analyzing repeated measures data using duplicateCorrelation.
# #The model forces the magnitude of the random effect to be the same across all genes.
# #Estimate linear mixed model with a single variance component
#vfit_c <- lmFit(vobj_c, design_c, block=d_n$samples$ID, correlation=dupcor_c$consensus)
#vfit_c <- contrasts.fit(vfit_c, contrasts=contr.matrix.c)
#
# #Save the model to avoid recomputing everything
# #save vfit_c
#saveRDS(vfit_c, file = here::here("Models","vfit_c.rds"))
#
#Load vfit_c
vfit_c <- readRDS(here::here("Models","vfit_c.rds"))

# Fit Empirical Bayes for moderated t-statistics, empirical Bayes moderation is carried out by borrowing information across all the genes to obtain more precise estimates of gene-wise variability
efit_c <- eBayes(vfit_c)
topTable(efit_c, n=20)
plotSA(efit_c, main="Final model: Mean-variance trend")
plotMD(efit_c)
abline(h=0,col="darkgrey")

#adjusted p-value cutoff that is set at 5% by default.
efit_c$F
#head(efit_c$F.p.value) #pval of the moderated F-statistics (equivalent to one-way ANOVA), it combines all contrast, so in this case it is the same as the t-statistic
#head(efit_c$p.value) #same
summary(decideTests(efit_c, adjust.method="fdr"))
#set log-fold-changes (log-FCs) to be above a minimum value. The treat method (McCarthy and Smyth 2009) used to calculate p-values from empirical Bayes moderated t-statistics with a minimum log-FC requirement. 
#no need to use the ebayes output for this

#Positive values means the gene is more expressed
tfit_c <- treat(vfit_c, lfc=lfc) 
topTreat(tfit_c)
nrow(tfit_c)
dt_c <- decideTests(tfit_c, adjust.method="fdr")
summary(dt_c) 

#Test for significance
Allgenes_c <- as.data.frame(topTreat(tfit_c, n=nrow(tfit_c)))
Filteredgenes_c <- subset(Allgenes_c, abs(logFC) >= lfc & adj.P.Val, 2 < adj.P.value)
nrow(subset(Filteredgenes_c, Filteredgenes_c$logFC > 0))
nrow(subset(Filteredgenes_c, Filteredgenes_c$logFC < 0))

#Significance
Allgenes_c$Significance <- "'Not significant'"
Allgenes_c$Significance[Allgenes_c$adj.P.Val < adj.P.value & Allgenes_c$logFC >= lfc] <- "Upregulated"
Allgenes_c$Significance[Allgenes_c$adj.P.Val < adj.P.value & Allgenes_c$logFC <= -lfc] <- "Downregulated"

#Plot DEG Volcano plot
c_plot <- ggplot(Allgenes_c, aes(x = logFC, y = -log10(P.Value), color = Significance, size = Significance)) +
  geom_point() +
  scale_color_manual(values = clrs) +  
  scale_size_manual(values = 0.5) +
  labs(title = "(d) Colony", x = "Log2 fold change", y = "-Log10 P-value") +
  theme_classic() +
  ylim(0,50) +
  xlim(-4,4) +
    theme(
    legend.position = "top",
    plot.title.position = "plot")
```

## Season

```{r}
#Tagged out to not run dupcor again (the saved outcome can be called from the dupcor folder)
# #Create a design matrix
# design_y <- model.matrix(~ -1 + year, data = d_n$samples) #model contrasts without intercept is easier
# colnames(design_y) <- c("Season1", "Season2") #clean group names
# #create matrix for contrasts
# contr.matrix.y <- makeContrasts(
#   Season1vsSeason2 = Season1-Season2,
#   levels = colnames(design_y)) 
#
# #voom() converts the read counts to log2-cpm, with associated weights, ready for linear modelling
# vobj_y <- voom(d_n, design_y, plot=TRUE)
# 
# #Fit a random effect
# dupcor_y <- duplicateCorrelation(vobj_y, design_y, block = d_n$samples$ID)
# dupcor_y$consensus #the estimated intra individual correlation (0.1744074)
# #the intra individual correlation will change the voom weights slightly
# #run voom again considering the duplicateCorrelation results in order to compute more accurate precision weights
# vobj_y = voom(d_n, design_y, plot=TRUE, 
#             block=d_n$samples$ID, correlation=dupcor_y$consensus)
# #update the correlation for the new voom weights
# dupcor_y <- duplicateCorrelation(vobj_y, design_y, block = d_n$samples$ID)
# dupcor_y$consensus #the estimated intra individual correlation (0.1744428)
#
# #Analyzing repeated measures data using duplicateCorrelation.
# #The model forces the magnitude of the random effect to be the same across all genes.
#
# #Estimate linear mixed model with a single variance component
# vfit_y <- lmFit(vobj_y, design_y, block=d_n$samples$ID, correlation=dupcor_y$consensus)
# vfit_y <- contrasts.fit(vfit_y, contrasts=contr.matrix.y)
#
# #Save the model to avoid recomputing everything
# saveRDS(vfit_y, file = here::here("Models","vfit_y.rds"))
#
#Load vfit_y
vfit_y <- readRDS(here::here("Models","vfit_y.rds"))
#ebayes model
efit_y <- eBayes(vfit_y)
topTable(efit_y, n=20)
plotSA(efit_y, main="Final model: Mean-variance trend")
plotMD(efit_y)
abline(h=0,col="darkgrey")

#adjusted p-value cutoff that is set at 5% by default.
efit_y$F
#head(efit_y$F.p.value) #pval of the moderated F-statistics (equivalent to one-way ANOVA), it combines all contrast, so in this case it is the same as the t-statistic
#head(efit_y$p.value) #same
summary(decideTests(efit_y, adjust.method="fdr"))
#set lfc 
tfit_y <- treat(vfit_y, lfc=lfc) 
topTreat(tfit_y)
nrow(tfit_y)
dt_y <- decideTests(tfit_y, adjust.method="fdr")
summary(dt_y) 

#Test for significance
Allgenes_y <- as.data.frame(topTreat(tfit_y, n=nrow(tfit_y)))
Filteredgenes_y <- subset(Allgenes_y, abs(logFC) >= lfc & adj.P.Val < adj.P.value)
nrow(subset(Filteredgenes_y, Filteredgenes_y$logFC > 0))
nrow(subset(Filteredgenes_y, Filteredgenes_y$logFC < 0))

#Plot the outcome
Allgenes_y$Significance <- "'Not significant'"
Allgenes_y$Significance[Allgenes_y$adj.P.Val < adj.P.value & Allgenes_y$logFC >= lfc] <- "Upregulated"
Allgenes_y$Significance[Allgenes_y$adj.P.Val < adj.P.value & Allgenes_y$logFC <= -lfc] <- "Downregulated"

#Plot DEG
y_plot <- ggplot(Allgenes_y, aes(x = logFC, y = -log10(P.Value), color = Significance, size = Significance)) +
  geom_point() +
  scale_color_manual(values = clrs) +  
  scale_size_manual(values = 0.5) +
  labs(title = "(c) Season", x = "Log2 fold change", y = "-Log10 P-value") +
  theme_classic() +
  ylim(0,50) +
  xlim(-4,4) +
    theme(
    legend.position = "top",
    plot.title.position = "plot")
```

```{r}
#Bind plots together
# Apply theme to make similar
y_plot <- y_plot + fig_theme
lhs_plot <- lhs_plot + fig_theme
c_plot <- c_plot + fig_theme
tp_plot <- tp_plot + fig_theme

# Combine plots
DEG_all <- ggarrange(tp_plot, lhs_plot, y_plot, c_plot, common.legend = T, legend = "bottom")

ggsave(here::here("Plots","DEG_all.png"), plot = DEG_all, width = 7.2, height = 5, dpi = 600)
```

# Interactions
## Year and life history stage interaction
```{r}
#Simple interaction (limma manual 9.5.2)
#Create a design matrix including life history stage and its interaction with year
d_n$samples$y.lhs <- as.factor(paste(d_n$samples$year, d_n$samples$life_history_stage, sep = "."))
design_interaction <- model.matrix(~ 0+y.lhs, data = d_n$samples)  # No intercept
#Clean column names
colnames(design_interaction) <- c("Season1.mum", "Season1.pup", "Season2.mum", "Season2.pup")

# #voom() converts the read counts to log2-cpm, with associated weights, ready for linear modelling
vobj_interaction <- voom(d_n, design_interaction, plot=TRUE)

#Fit a random effect
dupcor_interaction <- duplicateCorrelation(vobj_interaction, design_interaction, block = d_n$samples$ID)
# #dupcor_interaction$consensus #the estimated intra individual correlation (0.09626666)

vobj_interaction = voom(d_n, design_interaction, plot=TRUE, 
                        block=d_n$samples$ID, correlation=dupcor_interaction$consensus)
# #update the correlation for the new voom weights
dupcor_interaction <- duplicateCorrelation(vobj_interaction, design_interaction, block = d_n$samples$ID)
# #dupcor_interaction$consensus #the estimated intra individual correlation (0.09627793)

#Estimate linear mixed model with a single variance component
vfit_interaction <- lmFit(vobj_interaction, design_interaction, block=d_n$samples$ID, correlation=dupcor_interaction$consensus)

#Fit the contrasts we are interested in
cont.matrix.interaction <- makeContrasts(
  SeasonalAdult = Season1.mum - Season2.mum, #which genes in adults are DE among seasons
  SeasonalPup = Season1.pup - Season2.pup,  #which genes in pups are DE among seasons
  Interaction = (Season1.mum - Season1.pup)-(Season2.mum - Season2.pup), #which genes respond differently in pups compared to adults 
  levels = colnames(design_interaction)
)

vfit_interaction <- contrasts.fit(vfit_interaction, contrasts=cont.matrix.interaction)
# Fit Empirical Bayes for moderated t-statistics, empirical Bayes moderation is carried out by borrowing information across all the genes to obtain more precise estimates of gene-wise variability
efit_interaction <- eBayes(vfit_interaction)
topTable(efit_interaction, n=20)
plotSA(efit_interaction, main="Final model: Mean-variance trend")
plotMD(efit_interaction)
abline(h=0,col="darkgrey")

#adjusted p-value cutoff that is set at 5% by default.
efit_interaction$F
#head(efit_y$F.p.value) #pval of the moderated F-statistics (equivalent to one-way ANOVA), it combines all contrast, so in this case it is the same as the t-statistic
#head(efit_y$p.value) #same
summary(decideTests(efit_interaction, adjust.method="fdr"))

tfit_interaction <- treat(vfit_interaction, lfc=lfc) 
topTreat(tfit_interaction)
nrow(tfit_interaction)
dt_interaction <- decideTests(tfit_interaction, adjust.method="fdr")
summary(dt_interaction) 
```
## Year and colony interaction

```{r}
#Simple interaction (limma manual 9.5.2)
#Create a design matrix including life history stage and its interaction with colony
d_n$samples$y.colony <- as.factor(paste(d_n$samples$year, d_n$samples$colony, sep = "."))
design_interaction2 <- model.matrix(~ 0+y.colony, data = d_n$samples)  # No intercept
#Clean column names
colnames(design_interaction2) <- c("Season1.FWB", "Season1.SSB", "Season2.FWB", "Season2.SSB")

# #voom() converts the read counts to log2-cpm, with associated weights, ready for linear modelling
vobj_interaction2 <- voom(d_n, design_interaction2, plot=TRUE)

#Fit a random effect
dupcor_interaction2 <- duplicateCorrelation(vobj_interaction2, design_interaction2, block = d_n$samples$ID)
# #dupcor_interaction$consensus #the estimated intra individual correlation (0.09626666)

vobj_interaction2 = voom(d_n, design_interaction2, plot=TRUE, 
                        block=d_n$samples$ID, correlation=dupcor_interaction2$consensus)
# #update the correlation for the new voom weights
dupcor_interaction2 <- duplicateCorrelation(vobj_interaction2, design_interaction2, block = d_n$samples$ID)
# #dupcor_interaction$consensus #the estimated intra individual correlation (0.09627793)

#Estimate linear mixed model with a single variance component
vfit_interaction2 <- lmFit(vobj_interaction2, design_interaction2, block=d_n$samples$ID, correlation=dupcor_interaction2$consensus)

#Fit the contrasts we are interested in
cont.matrix.interaction2 <- makeContrasts(
  FWBvsSSBseason1 = Season1.FWB - Season1.SSB, #which genes in season 1 are DE among colonies
  FWBvsSSBseason2 = Season2.FWB - Season2.SSB, #which genes in season 2 are DE among colonies
  Interaction = (Season1.FWB - Season1.SSB)-(Season2.FWB - Season2.SSB), #which genes respond differently in SSB compared to FWB
  levels = colnames(design_interaction2)
)

vfit_interaction2 <- contrasts.fit(vfit_interaction2, contrasts=cont.matrix.interaction2)
# Fit Empirical Bayes for moderated t-statistics, empirical Bayes moderation is carried out by borrowing information across all the genes to obtain more precise estimates of gene-wise variability
efit_interaction2 <- eBayes(vfit_interaction2)
topTable(efit_interaction2, n=20)
plotSA(efit_interaction2, main="Final model: Mean-variance trend")
plotMD(efit_interaction2)
abline(h=0,col="darkgrey")

#adjusted p-value cutoff that is set at 5% by default.
efit_interaction2$F
#head(efit_y$F.p.value) #pval of the moderated F-statistics (equivalent to one-way ANOVA), it combines all contrast, so in this case it is the same as the t-statistic
#head(efit_y$p.value) #same
summary(decideTests(efit_interaction2, adjust.method="fdr"))

tfit_interaction2 <- treat(vfit_interaction2, lfc=lfc) 
topTreat(tfit_interaction2)
nrow(tfit_interaction2)
dt_interaction2 <- decideTests(tfit_interaction2, adjust.method="fdr")
summary(dt_interaction2) 
```

## Time point and life stage interaction

```{r}
#Simple interaction (limma manual 9.5.2)
#Create a design matrix including life history stage and its interaction with colony
d_n$samples$tp.lhs <- as.factor(paste(d_n$samples$group, d_n$samples$life_history_stage, sep = "."))
design_interaction3 <- model.matrix(~ 0+tp.lhs, data = d_n$samples)  # No intercept
#Clean column names
colnames(design_interaction3) <- c("Start.Mum", "Start.Pup", "End.Mum", "End.Pup")

# #voom() converts the read counts to log2-cpm, with associated weights, ready for linear modelling
vobj_interaction3 <- voom(d_n, design_interaction3, plot=TRUE)

#Fit a random effect
dupcor_interaction3 <- duplicateCorrelation(vobj_interaction3, design_interaction3, block = d_n$samples$ID)
dupcor_interaction3$consensus #the estimated intra individual correlation (0.2608585)

vobj_interaction3 = voom(d_n, design_interaction3, plot=TRUE, 
                        block=d_n$samples$ID, correlation=dupcor_interaction3$consensus)
# #update the correlation for the new voom weights
dupcor_interaction3 <- duplicateCorrelation(vobj_interaction3, design_interaction3, block = d_n$samples$ID)
dupcor_interaction3$consensus #the estimated intra individual correlation (0.2909559)

#Estimate linear mixed model with a single variance component
vfit_interaction3 <- lmFit(vobj_interaction3, design_interaction3, block=d_n$samples$ID, correlation=dupcor_interaction3$consensus)

#Fit the contrasts we are interested in
cont.matrix.interaction3 <- makeContrasts(
  EndvsStartMum = End.Mum - Start.Mum, #which genes in mothers are DE among tp
  EndvsStartPup = End.Pup - Start.Pup,  #which genes in pups are DE among tp
  Interaction = (End.Mum - End.Pup)-(Start.Mum - Start.Pup), #which genes respond differently over time in mums compared to pups
  levels = colnames(design_interaction3)
)

vfit_interaction3 <- contrasts.fit(vfit_interaction3, contrasts=cont.matrix.interaction3)
# Fit Empirical Bayes for moderated t-statistics, empirical Bayes moderation is carried out by borrowing information across all the genes to obtain more precise estimates of gene-wise variability
efit_interaction3 <- eBayes(vfit_interaction3)
topTable(efit_interaction3, n=20)
plotSA(efit_interaction3, main="Final model: Mean-variance trend")
plotMD(efit_interaction3)
abline(h=0,col="darkgrey")

#adjusted p-value cutoff that is set at 5% by default.
efit_interaction3$F
#head(efit_interaction3$F.p.value) #pval of the moderated F-statistics (equivalent to one-way ANOVA), it combines all contrast, so in this case it is the same as the t-statistic
#head(efit_interaction3$p.value) #same (not the same?)
summary(decideTests(efit_interaction3, adjust.method="fdr"))

tfit_interaction3 <- treat(vfit_interaction3, lfc=lfc) 
topTreat(tfit_interaction3)
topTreat(tfit_interaction3, coef = "Interaction") #specific for the interaction
nrow(tfit_interaction3)
dt_interaction3 <- decideTests(tfit_interaction3, adjust.method="fdr")
summary(dt_interaction3) 

#Test for significance
#Extract significant genes for each contrast
Allgenes_EndvsStartMum <- as.data.frame(topTreat(tfit_interaction3, coef = "EndvsStartMum", n = nrow(tfit_interaction3)))
Allgenes_EndvsStartPup <- as.data.frame(topTreat(tfit_interaction3, coef = "EndvsStartPup", n = nrow(tfit_interaction3)))
Allgenes_Interaction3 <- as.data.frame(topTreat(tfit_interaction3, coef = "Interaction", n = nrow(tfit_interaction3)))
#Filtered
Filtered_EndvsStartMum <- subset(Allgenes_EndvsStartMum, abs(logFC) >= lfc & adj.P.Val < adj.P.value)
Filtered_EndvsStartPup <- subset(Allgenes_EndvsStartPup, abs(logFC) >= lfc & adj.P.Val < adj.P.value)
Filtered_Interaction3 <- subset(Allgenes_Interaction3, abs(logFC) >= lfc & adj.P.Val < adj.P.value)

Upregulated_EndvsStartMum <- subset(Filtered_EndvsStartMum, logFC > 0)
Downregulated_EndvsStartMum <- subset(Filtered_EndvsStartMum, logFC < 0)

Upregulated_EndvsStartPup <- subset(Filtered_EndvsStartPup, logFC > 0)
Downregulated_EndvsStartPup <- subset(Filtered_EndvsStartPup, logFC < 0)

Upregulated_Interaction3 <- subset(Filtered_Interaction3, logFC > 0)
Downregulated_Interaction3 <- subset(Filtered_Interaction3, logFC < 0)

# Create a list to save the up and down regulated genes and the df containing all genes evaluated for GO analysis
interaction3_list <- list(All_Interaction3 = Allgenes_Interaction3, 
                          Upregulated_interaction3 = Upregulated_Interaction3, 
                          Downregulated_Interaction3 = Downregulated_Interaction3, 
                          Upregulated_EndvsStartPup = Upregulated_EndvsStartPup,
                          Downregulated_EndvsStartPup = Downregulated_EndvsStartPup,
                          Upregulated_EndvsStartMum = Upregulated_EndvsStartMum,
                          Downregulated_EndvsStartMum = Downregulated_EndvsStartMum)

# Loop through and save each data frame using `here()`
for (name in names(interaction3_list)) {
  write.csv(interaction3_list[[name]], file = here::here("GenesGO", paste0(name, ".csv")), row.names = TRUE)
}

#Gene check
#Names <- c("kiaa1522", "fos", "abcc8", "dppa4", "ddit4", "ildr2", "sipa1l2", "arnt2", "crybg2", "gucy2c", "mast4", "pitpnm3", "slc4a8", "tmem150c", "aunip", "chrm3", "mybl1", "kctd19", "ciart")
#Allgenes_EndvsStartMum[Names,]
#Allgenes_EndvsStartPup[Names,]
#Allgenes_Interaction3[Names,]
```
## Colony and life stage interaction

```{r}
#Simple interaction (limma manual 9.5.2)
#Create a design matrix including life history stage and its interaction with colony
d_n$samples$colony.lhs <- as.factor(paste(d_n$samples$colony, d_n$samples$life_history_stage, sep = "."))
design_interaction4 <- model.matrix(~ 0+colony.lhs, data = d_n$samples)  # No intercept
#Clean column names
colnames(design_interaction4) <- c("FWB.Mum", "FWB.Pup", "SSB.Mum", "SSB.Pup")

# #voom() converts the read counts to log2-cpm, with associated weights, ready for linear modelling
vobj_interaction4 <- voom(d_n, design_interaction4, plot=TRUE)

#Fit a random effect
dupcor_interaction4 <- duplicateCorrelation(vobj_interaction4, design_interaction4, block = d_n$samples$ID)
# #dupcor_interaction4$consensus #the estimated intra individual correlation (0.1308229)

vobj_interaction4 = voom(d_n, design_interaction4, plot=TRUE, 
                        block=d_n$samples$ID, correlation=dupcor_interaction4$consensus)
# #update the correlation for the new voom weights
dupcor_interaction4 <- duplicateCorrelation(vobj_interaction4, design_interaction4, block = d_n$samples$ID)
# #dupcor_interaction4$consensus #the estimated intra individual correlation (0.1308384)

#Estimate linear mixed model with a single variance component
vfit_interaction4 <- lmFit(vobj_interaction4, design_interaction4, block=d_n$samples$ID, correlation=dupcor_interaction4$consensus)

#Fit the contrasts we are interested in
cont.matrix.interaction4 <- makeContrasts(
  SSBvsFWBMum = SSB.Mum - FWB.Mum, #which genes in mothers are DE among colonies
  SSBvsFWBPup = SSB.Pup - FWB.Pup,  #which genes in pups are DE among colonies
  Interaction = (SSB.Mum - SSB.Pup)-(FWB.Mum - FWB.Pup), #which genes respond different in the two colonies in mums compared to pups
  levels = colnames(design_interaction4)
)

vfit_interaction4 <- contrasts.fit(vfit_interaction4, contrasts=cont.matrix.interaction4)
# Fit Empirical Bayes for moderated t-statistics, empirical Bayes moderation is carried out by borrowing information across all the genes to obtain more precise estimates of gene-wise variability
efit_interaction4 <- eBayes(vfit_interaction4)
topTable(efit_interaction4, n=20)
plotSA(efit_interaction4, main="Final model: Mean-variance trend")
plotMD(efit_interaction4)
abline(h=0,col="darkgrey")

#adjusted p-value cutoff that is set at 5% by default.
efit_interaction4$F
#head(efit_interaction4$F.p.value) #pval of the moderated F-statistics (equivalent to one-way ANOVA), it combines all contrast, so in this case it is the same as the t-statistic
#head(efit_interaction4$p.value) #same (not the same?)
summary(decideTests(efit_interaction4, adjust.method="fdr"))

tfit_interaction4 <- treat(vfit_interaction4, lfc=lfc) 
topTreat(tfit_interaction4)
topTreat(tfit_interaction4, coef = "Interaction") #specific for the interaction
nrow(tfit_interaction4)
dt_interaction4 <- decideTests(tfit_interaction4, adjust.method="fdr")
summary(dt_interaction4) 
```

# GOterms using clusterProfiler

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("enrichplot")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("clusterProfiler")

invisible(pacman::p_load(dplyr, clusterProfiler, org.Hs.eg.db, enrichplot, ggplot2, VennDiagram, ggrepel))
```

## Time point

```{r}
#Call up or down regulated genes from DEG in TP and LHS
#t& transform genesymbols to upper case
Up_TP <- read.csv("GenesGO/Upregulated_tp.csv") %>%
  mutate(X = toupper(X))
colnames(Up_TP)[1] <- "GeneSymbol"
Down_TP <- read.csv("GenesGO/Downregulated_tp.csv") %>%
  mutate(X = toupper(X))
colnames(Down_TP)[1] <- "GeneSymbol"
TP <- rbind(Up_TP, Down_TP)

#background list
Background <- read.csv("GenesGO/All_tp.csv") %>%
  mutate(X = toupper(X))
Background <- Background[1]
colnames(Background) <- "GeneSymbol"

```

```{r}
#Go enrichment analysis for up and down regulated genes
#Go enrichment analysis
GOterms_TP_UP <- enrichGO(gene = Up_TP$GeneSymbol, OrgDb = org.Hs.eg.db, universe = Background$GeneSymbol, keyType = "SYMBOL", ont = "BP", pvalueCutoff = 0.05, pAdjustMethod = "BH")

#simplify/group gene list as maybe revigo?

GOterms_TP_UP_grouped<- simplify(
  GOterms_TP_UP,
  cutoff = 0.5,              # Similarity threshold (lower = stricter)
  by = "p.adjust",           # Keep the more significant term
  select_fun = min,          # Function to select representative term
  measure = "Wang"           # Semantic similarity measure
)

GOterms_TP_UP <- as.data.frame(GOterms_TP_UP)
GOterms_TP_UP$DEG <- "Upregulated"



GOterms_TP_Down <- enrichGO(gene = Down_TP$GeneSymbol, OrgDb = org.Hs.eg.db, universe = Background$GeneSymbol, keyType = "SYMBOL", ont = "BP", pvalueCutoff = 0.05, pAdjustMethod = "BH")

#simplify/group gene list as maybe revigo?

GOterms_TP_Down_grouped<- simplify(
  GOterms_TP_Down,
  cutoff = 0.5,              # Similarity threshold (lower = stricter)
  by = "p.adjust",           # Keep the more significant term
  select_fun = min,          # Function to select representative term
  measure = "Wang"           # Semantic similarity measure
)

GOterms_TP_Down <- as.data.frame(GOterms_TP_Down)
GOterms_TP_Down$DEG <- "Downregulated"

GOterms_TP <- rbind(GOterms_TP_UP, GOterms_TP_Down)
```

### VennDiagram
```{r}
myCol <- c("#9D00FF", "#FFA500")

grid.newpage()  # Clears the plot area

draw.pairwise.venn(
  area1 = length(GOterms_TP_UP$ID),
  area2 = length(GOterms_TP_Down$ID),
  cross.area = length(intersect(GOterms_TP_UP$ID, GOterms_TP_Down$ID)),
  category = c("Upregulated", "Downregulated"),
  cex = 1,
  cat.cex = 1.2,
  height = 480, 
  width = 480, 
  
  #Circles
  lwd = 2,
  lty = 'blank',
  col= myCol,
  fill = c(alpha("#9D00FF",0.3), alpha('#FFA500',0.3)),
  
  # Numbers
 #fontface = "bold",
  fontfamily = "sans",
 
  cat.pos = c(-20, 20),      
  cat.dist = c(0.03, 0.03),  
  cat.default.pos = "text")
```



```{r}
#Go enrichment analysis

top20_labels <- GOterms_TP %>%
  arrange(p.adjust) %>%
  slice_head(n = 20)

#Bubble plot fancy

ggplot(GOterms_TP, aes(x = zScore, y = -log10(p.adjust))) +
  geom_point(aes(size = Count, fill = Description), shape = 24, color = "black", alpha = 0.7) +
  scale_size_continuous(name = "DEG Count") +
  labs(
    title = "Significantly Enriched GO Terms for DEGs",
    x = "Z-score",
    y = expression(-log[10]("Adjusted p-value (FDR)"))
  ) +
  theme_classic() +
  theme(legend.position = "right") +
  guides(fill = "none") 


#Fancy smancy bubble plot
ggplot(GOterms_TP, aes(x = zScore, y = -log10(p.adjust))) +
  geom_point(aes(size = Count, shape = DEG, fill = Description), color = "black", alpha = 0.7) +
  
  scale_shape_manual(values = c("Upregulated" = 24, "Downregulated" = 25)) +  # circle & triangle
  scale_size_continuous(name = "DEG Count") +
  
geom_text_repel(
  data = top20_labels,  # your top terms
  aes(label = ID),
  nudge_y = 0.1,         # nudges label upward
  size = 3,
  segment.color = NA,    #remove line
  max.overlaps = Inf
) +
  
  labs(
    title = "Significantly Enriched GO Terms for DEGs",
    x = "Z-score",
    y = expression(-log[10]("Adjusted p-value (FDR)")),
    shape = "Regulation"
  ) +
  
  theme_classic() +
  theme(legend.position = "right") +
  guides(
  size = guide_legend(
    override.aes = list(shape = 24)  # use triangle in size legend
  ),
  fill = "none" 
)
```


```{r}
#Go enrichment analysis
GO_results_LHS <- enrichGO(gene = LHS$GeneSymbol, OrgDb = org.Hs.eg.db, universe = Background$GeneSymbol, keyType = "SYMBOL", ont = "BP", pvalueCutoff = 0.05, pAdjustMethod = "BH")

as.data.frame(GO_results_LHS)
#Bubble plot
dotplot(GO_results_LHS, showCategory = 20) +
  ggtitle("Top Enriched GO Biological Processes")

barplot(GO_results_LHS, showCategory = 20) +
  ggtitle("Top Enriched GO Biological Processes")

go_df_lhs <- as.data.frame(GO_results_LHS)

ggplot(go_df_lhs, aes(x = zScore, y = -log10(p.adjust))) +
  geom_point(aes(size = Count, fill = Description), shape = 21, color = "black", alpha = 0.7) +
  scale_size_continuous(name = "DEG Count") +
  labs(
    title = "Significantly Enriched GO Terms for DEGs",
    x = "Z-score",
    y = expression(-log[10]("Adjusted p-value (FDR)"))
  ) +
  theme_classic() +
  theme(legend.position = "right") +
  guides(fill = "none") 

#simplify/group gene list as maybe revigo?

go_df_lhs_grouped <- simplify(
  GO_results_LHS,
  cutoff = 0.5,              # Similarity threshold (lower = stricter)
  by = "p.adjust",           # Keep the more significant term
  select_fun = min,          # Function to select representative term
  measure = "Wang"           # Semantic similarity measure
)

go_df_lhs_grouped <- as.data.frame(go_df_lhs_grouped)

ggplot(go_df_lhs_grouped , aes(x = zScore, y = -log10(p.adjust))) +
  geom_point(aes(size = Count, fill = Description), shape = 21, color = "black", alpha = 0.7) +
  geom_text(aes(label = Description), size = 3, vjust = -1) +
  scale_size_continuous(name = "DEG Count") +
  labs(
    title = "Significantly Enriched GO Terms for DEGs",
    x = "Z-score",
    y = expression(-log[10]("Adjusted p-value (FDR)"))
  ) +
  theme_classic() +
  theme(legend.position = "right") +
  guides(fill = "none") 

go_df_lhs$ID
```

### Revigo in R
```{r}
remotes::install_github("ssayols/rrvgo")
library(rrvgo)
```

### Similarity matrix
```{r}
simMatrix <- calculateSimMatrix(GOterms_TP_UP$ID,
                                orgdb="org.Hs.eg.db",
                                ont="BP",
                                method="Rel")

scores <- setNames(-log10(GOterms_TP_UP$qvalue), GOterms_TP_UP$ID)
reducedTerms <- reduceSimMatrix(simMatrix,
                                scores,
                                threshold=0.7,
                                orgdb="org.Hs.eg.db")
heatmapPlot(simMatrix,
            reducedTerms,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)

#Same but with GOterms from GOrilla
GOrilla_TP_Down <- c("GO:0072376", "GO:0009617" , "GO:0051707", "GO:0006952", "GO:0045959" , "GO:0030450", "GO:0006959", "GO:0043207", "GO:0006958", "GO:1903027", "GO:0009608", "GO:0009609", "GO:0009607", "GO:0006956", "GO:0030449", "GO:2000257", "GO:0002924", "GO:0002376", "GO:0051704", "GO:0002252", "GO:2000258", "GO:0045916", "GO:0002920", "GO:0002921", "GO:0002713", "GO:0002890", "GO:0002923", "GO:0002673", "GO:0045055", "GO:0070613", "GO:1903317", "GO:0030162", "GO:0006887", "GO:0009605", "GO:0018149", "GO:0032940", "GO:0042742", "GO:0045087", "GO:1903318", "GO:0010955", "GO:0046903", "GO:0002707", "GO:0002704", "GO:0045732", "GO:0002823", "GO:0002274", "GO:0002820", "GO:0030100", "GO:0002712", "GO:0019730", "GO:0002889", "GO:0002215", "GO:1905892", "GO:1905891", "GO:1905895", "GO:1905894", "GO:0072139", "GO:0072378", "GO:2000768", "GO:1902996", "GO:1902998", "GO:1902847", "GO:0007620", "GO:0045861")

GOrilla_TP_Down_qvalue <- c(0.00273, 0.00294, 0.00307, 0.00358, 0.00693, 0.00577, 0.00885, 0.00847, 0.00759, 0.00693, 0.00630, 0.00577, 0.00638, 0.00888, 0.00829, 0.00831, 0.01020, 0.01120, 0.01140, 0.01150, 0.01150, 0.01100, 0.01360, 0.02160, 0.02530, 0.02430, 0.02340, 0.02430, 0.02410, 0.02630, 0.02730, 0.03540, 0.03560, 0.04560, 0.04450, 0.07400, 0.07740, 0.07800, 0.08080, 0.07880, 0.10500, 0.12600, 0.16600, 0.17200, 0.16800, 0.17000, 0.19800, 0.20900, 0.22800, 0.22400, 0.21900, 0.25800, 0.25300, 0.24900, 0.24400, 0.24000, 0.23500, 0.23100, 0.22700, 0.22400, 0.22000, 0.21600, 0.21300, 0.21300)

GOrilla_TP_Down <- as.data.frame(cbind(GOrilla_TP_Down, GOrilla_TP_Down_qvalue))
GOrilla_TP_Down <- subset(GOrilla_TP_Down, GOrilla_TP_Down_qvalue < 0.05)

simMatrix_GO <- calculateSimMatrix(GOrilla_TP_Down$GOrilla_TP_Down,
                                orgdb="org.Hs.eg.db",
                                ont="BP",
                                method="Rel")

scores_GO <- setNames(-log10(as.numeric(GOrilla_TP_Down$GOrilla_TP_Down_qvalue)), GOrilla_TP_Down$GOrilla_TP_Down)
reducedTerms_GP <- reduceSimMatrix(simMatrix_GO,
                                scores_GO,
                                threshold=0.7,
                                orgdb="org.Hs.eg.db")
heatmapPlot(simMatrix_GO,
            reducedTerms_GP,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```
### Scatterplot of GOTerms
```{r}
R_GO <- scatterPlot(simMatrix, reducedTerms) + theme_classic() + 
  ggtitle("Semantic Similarity of GO Terms: Downregulated TP2") +
  xlim(-0.6, 0.4) +
  ylim(-0.4,0.6)

GOrilla_GO <- scatterPlot(simMatrix_GO, reducedTerms_GP) + theme_classic() + 
  ggtitle("Semantic Similarity of GO Terms: GOrilla") +
  xlim(-0.6, 0.4) +
  ylim(-0.4,0.6)

#library(cowplot)

plot_grid(R_GO, GOrilla_GO)
```

### Similarity matrix
```{r}
simMatrix_up <- calculateSimMatrix(GOterms_TP_Down$ID,
                                orgdb="org.Hs.eg.db",
                                ont="BP",
                                method="Rel")

scores_up <- setNames(-log10(GOterms_TP_Down$qvalue), GOterms_TP_Down$ID)
reducedTerms_up <- reduceSimMatrix(simMatrix_up,
                                scores_up,
                                threshold=0.7,
                                orgdb="org.Hs.eg.db")
heatmapPlot(simMatrix_up,
            reducedTerms_up,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```
### Scatterplot of GOTerms
```{r}
R_GO_Up <- scatterPlot(simMatrix_up, reducedTerms_up) + theme_classic() + 
  ggtitle("Semantic Similarity of GO Terms: Upregulated TP2") 

plot_grid(R_GO, R_GO_Up)
```
## Life history stage

```{r}
#lhs
Up_LHS <- read.csv("GenesGO/Upregulated_lhs.csv") %>%
  mutate(X = toupper(X))
colnames(Up_LHS)[1] <- "GeneSymbol"
Down_LHS <- read.csv("GenesGO/Downregulated_lhs.csv") %>%
  mutate(X = toupper(X))
colnames(Down_LHS)[1] <- "GeneSymbol"
LHS <- rbind(Up_LHS, Down_LHS)
```

```{r}
#Go enrichment analysis
GOterms_LHS_UP <- enrichGO(gene = Up_LHS$GeneSymbol, OrgDb = org.Hs.eg.db, universe = Background$GeneSymbol, keyType = "SYMBOL", ont = "BP", pvalueCutoff = 0.05, pAdjustMethod = "BH")
```


### Revigo
```{r}
simMatrix_LHS_UP <- calculateSimMatrix(GOterms_LHS_UP$ID,
                                orgdb="org.Hs.eg.db",
                                ont="BP",
                                method="Rel")

scores_LHS_UP <- setNames(-log10(GOterms_LHS_UP$qvalue), GOterms_LHS_UP$ID)
reducedTerms_LHS_UP <- reduceSimMatrix(simMatrix_LHS_UP,
                                scores_LHS_UP,
                                threshold=0.7,
                                orgdb="org.Hs.eg.db")
heatmapPlot(simMatrix_LHS_UP,
            reducedTerms_LHS_UP,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```
#Scatterplot
```{r}
R_GO_LHS_UP <- scatterPlot(simMatrix_LHS_UP, reducedTerms_LHS_UP) + theme_classic() + 
  ggtitle("Semantic Similarity of GO Terms: Downregulated TP2") 
```

```{r}
#Go enrichment analysis
GOterms_LHS_DOWN <- enrichGO(gene = Down_LHS$GeneSymbol, OrgDb = org.Hs.eg.db, universe = Background$GeneSymbol, keyType = "SYMBOL", ont = "BP", pvalueCutoff = 0.05, pAdjustMethod = "BH")
```


### Revigo
```{r}
simMatrix_LHS_DOWN <- calculateSimMatrix(GOterms_LHS_DOWN$ID,
                                orgdb="org.Hs.eg.db",
                                ont="BP",
                                method="Rel")

scores_LHS_DOWN <- setNames(-log10(GOterms_LHS_DOWN$qvalue), GOterms_LHS_DOWN$ID)
reducedTerms_LHS_DOWN <- reduceSimMatrix(simMatrix_LHS_DOWN,
                                scores_LHS_DOWN,
                                threshold=0.7,
                                orgdb="org.Hs.eg.db")
heatmapPlot(simMatrix_LHS_DOWN,
            reducedTerms_LHS_DOWN,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```
# Scatterplot
```{r}
R_GO_LHS_DOWN <- scatterPlot(simMatrix_LHS_DOWN, reducedTerms_LHS_DOWN) + theme_classic() + 
  ggtitle("Semantic Similarity of GO Terms: Downregulated TP2") 
```

# GOterms from GOrilla

```{r}
#LOad in GOterms from GOrilla
```

