---
title: "Gene expression: interaction"
subtitle: "Analysis script"
author: "Bernice Sepers & Ane Liv Berthelsen"

format:
  #html: 
  pdf: 
    papersize: "a4paper"
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: FALSE
    documentclass: article
    geometry:
      - inner=3cm
      - outer=3cm
      - top=2cm
      - bottom=3cm
      - headsep=22pt
      - headheight=11pt
      - footskip=33pt
      - ignorehead
      - ignorefoot
      - heightrounded
    fig-pos: 'h'
    fig-cap-location: bottom
    tbl-cap-location: top
editor: visual
date: 03/09/2025
mainfont: Times New Roman
---

```{r}
#install.packages("here")
#install.packages("pacman")
library(here)
here::here()
```

```{r}
#| label: packages
#| echo: FALSE
#here::here()
invisible(pacman::p_load(dplyr, readxl, tidyverse, cowplot, edgeR, DESeq2, RColorBrewer, ggplot2, reshape2, ggpubr, extrafont, pheatmap))

#font_import()
#fonts()
#cowplot for toggling with plots
#edgeR  loads limma as a dependency, limma for analysing microassay and RNA seq data
#DESeq2 needed for rowCounts
#reshape2 needed for the first plot
#ggpubr need to bind plots together
#extrafont for arial font
#pheatmap needed for heatmaps
```

# Data

```{r}
#| label: load_data
#| echo: FALSE
#Load in the filtered data
#Mutate all needed predictors into factors
d_n<-readRDS("rnaseq_filtered_Liv.Rdata")  
d_n$samples$life_history_stage <- as.factor(d_n$samples$life_history_stage)
d_n$samples$colony <- as.factor(d_n$samples$colony)
d_n$samples$ID <- as.factor(d_n$samples$ID)
d_n$samples$year <- as.factor(d_n$samples$year)
d_n$samples$group <- as.factor(d_n$samples$group)
d_n$samples$pair <- as.factor(d_n$samples$pair)
#dim(d_n) 12639 rows
#remove outlier
d_n <- d_n[, d_n$samples$Sample != "F6_FWB_mum_start_2018B"]
```

```{r}
#| label: color_scheme
#| echo: FALSE
clrs <- c("Upregulated" = "#FDA172", "Downregulated" = "#9867B5", "Not significant" = "grey20")
lfc <- 1 
adj.P.value <- 0.05
fig_theme <- theme(text = element_text(family = "Arial", size = 12),
                   legend.title = element_blank(),
                   legend.text=element_text(size=10),
                   axis.text.x = element_text(size =10),
                   axis.text.y = element_text(size = 10))

ann_colors <- list(
  TP = c("Day 0" = "#8B9E58", "Day 60" = "#638B9B"),
  LHS = c(Mother = "#C44B4E", Pup = "#C46A00")
)
```

# Differential gene expression analysis 
# Interactions
In this section, we perform a differential gene expression analysis between biologically relevant interactions between the main contrasts. 
## Year and life history stage interaction
Framework for differential gene expression analysis using interactions by ALB. 
```{r}
#| label: load_data
#| echo: FALSE
# #Simple interaction (limma manual 9.5.2)
# #Create a design matrix including life history stage and its interaction with year
# d_n$samples$y.lhs <- as.factor(paste(d_n$samples$year, d_n$samples$life_history_stage, sep = "."))
# design_interaction <- model.matrix(~ 0+y.lhs, data = d_n$samples)  # No intercept
# #Clean column names
# colnames(design_interaction) <- c("Season1.mum", "Season1.pup", "Season2.mum", "Season2.pup")
# 
# # #voom() converts the read counts to log2-cpm, with associated weights, ready for linear modelling
# vobj_interaction <- voom(d_n, design_interaction, plot=TRUE)
# 
# #Fit a random effect
# dupcor_interaction <- duplicateCorrelation(vobj_interaction, design_interaction, block = d_n$samples$ID)
# # #dupcor_interaction$consensus #the estimated intra individual correlation (0.09626666)
# 
# vobj_interaction = voom(d_n, design_interaction, plot=TRUE, 
#                         block=d_n$samples$ID, correlation=dupcor_interaction$consensus)
# # #update the correlation for the new voom weights
# dupcor_interaction <- duplicateCorrelation(vobj_interaction, design_interaction, block = d_n$samples$ID)
# # #dupcor_interaction$consensus #the estimated intra individual correlation (0.09627793)
# 
# #Estimate linear mixed model with a single variance component
# vfit_interaction <- lmFit(vobj_interaction, design_interaction, block=d_n$samples$ID, correlation=dupcor_interaction$consensus)
# 
# #Fit the contrasts we are interested in
# cont.matrix.interaction <- makeContrasts(
#   SeasonalAdult = Season1.mum - Season2.mum, #which genes in adults are DE among seasons
#   SeasonalPup = Season1.pup - Season2.pup,  #which genes in pups are DE among seasons
#   Interaction = (Season1.mum - Season1.pup)-(Season2.mum - Season2.pup), #which genes respond differently in pups compared to adults 
#   levels = colnames(design_interaction)
# )
# 
# vfit_interaction <- contrasts.fit(vfit_interaction, contrasts=cont.matrix.interaction)
#
# #Save the model to avoid recomputing everything
# saveRDS(vfit_interaction, file = here::here("Models","vfit_interaction.rds"))
#
#Load vfit_interaction
vfit_interaction <- readRDS(here::here("Models","vfit_interaction.rds"))

# Fit Empirical Bayes for moderated t-statistics, empirical Bayes moderation is carried out by borrowing information across all the genes to obtain more precise estimates of gene-wise variability
efit_interaction <- eBayes(vfit_interaction)
topTable(efit_interaction, n=20)
plotSA(efit_interaction, main="Final model: Mean-variance trend")
plotMD(efit_interaction)
abline(h=0,col="darkgrey")

#adjusted p-value cutoff that is set at 5% by default.
efit_interaction$F
#head(efit_y$F.p.value) #pval of the moderated F-statistics (equivalent to one-way ANOVA), it combines all contrast, so in this case it is the same as the t-statistic
#head(efit_y$p.value) #same
summary(decideTests(efit_interaction, adjust.method="fdr"))

tfit_interaction <- treat(vfit_interaction, lfc=lfc) 
topTreat(tfit_interaction)
nrow(tfit_interaction)
dt_interaction <- decideTests(tfit_interaction, adjust.method="fdr")
summary(dt_interaction) 
```

## Year and colony interaction
Framework for differential gene expression analysis using interactions by ALB. 
```{r}
#| label: load_data
#| echo: FALSE
# #Simple interaction (limma manual 9.5.2)
# #Create a design matrix including life history stage and its interaction with colony
# d_n$samples$y.colony <- as.factor(paste(d_n$samples$year, d_n$samples$colony, sep = "."))
# design_interaction2 <- model.matrix(~ 0+y.colony, data = d_n$samples)  # No intercept
# #Clean column names
# colnames(design_interaction2) <- c("Season1.FWB", "Season1.SSB", "Season2.FWB", "Season2.SSB")
# 
# # #voom() converts the read counts to log2-cpm, with associated weights, ready for linear modelling
# vobj_interaction2 <- voom(d_n, design_interaction2, plot=TRUE)
# 
# #Fit a random effect
# dupcor_interaction2 <- duplicateCorrelation(vobj_interaction2, design_interaction2, block = d_n$samples$ID)
# # #dupcor_interaction$consensus #the estimated intra individual correlation (0.09626666)
# 
# vobj_interaction2 = voom(d_n, design_interaction2, plot=TRUE, 
#                         block=d_n$samples$ID, correlation=dupcor_interaction2$consensus)
# # #update the correlation for the new voom weights
# dupcor_interaction2 <- duplicateCorrelation(vobj_interaction2, design_interaction2, block = d_n$samples$ID)
# # #dupcor_interaction$consensus #the estimated intra individual correlation (0.09627793)
# 
# #Estimate linear mixed model with a single variance component
# vfit_interaction2 <- lmFit(vobj_interaction2, design_interaction2, block=d_n$samples$ID, correlation=dupcor_interaction2$consensus)
# 
# #Fit the contrasts we are interested in
# cont.matrix.interaction2 <- makeContrasts(
#   FWBvsSSBseason1 = Season1.FWB - Season1.SSB, #which genes in season 1 are DE among colonies
#   FWBvsSSBseason2 = Season2.FWB - Season2.SSB, #which genes in season 2 are DE among colonies
#   Interaction = (Season1.FWB - Season1.SSB)-(Season2.FWB - Season2.SSB), #which genes respond differently in SSB compared to FWB
#   levels = colnames(design_interaction2)
# )
# 
# vfit_interaction2 <- contrasts.fit(vfit_interaction2, contrasts=cont.matrix.interaction2)
# 
# #Save the model to avoid recomputing everything
# saveRDS(vfit_interaction2, file = here::here("Models","vfit_interaction2.rds"))

#Load vfit_interaction
vfit_interaction2 <- readRDS(here::here("Models","vfit_interaction2.rds"))
# Fit Empirical Bayes for moderated t-statistics, empirical Bayes moderation is carried out by borrowing information across all the genes to obtain more precise estimates of gene-wise variability
efit_interaction2 <- eBayes(vfit_interaction2)
topTable(efit_interaction2, n=20)
plotSA(efit_interaction2, main="Final model: Mean-variance trend")
plotMD(efit_interaction2)
abline(h=0,col="darkgrey")

#adjusted p-value cutoff that is set at 5% by default.
efit_interaction2$F
#head(efit_y$F.p.value) #pval of the moderated F-statistics (equivalent to one-way ANOVA), it combines all contrast, so in this case it is the same as the t-statistic
#head(efit_y$p.value) #same
summary(decideTests(efit_interaction2, adjust.method="fdr"))

tfit_interaction2 <- treat(vfit_interaction2, lfc=lfc) 
topTreat(tfit_interaction2)
nrow(tfit_interaction2)
dt_interaction2 <- decideTests(tfit_interaction2, adjust.method="fdr")
summary(dt_interaction2) 
```

## Time point and life stage interaction
Framework for differential gene expression analysis using interactions by ALB. 
```{r}
#| label: load_data
#| echo: FALSE
# #Simple interaction (limma manual 9.5.2)
# #Create a design matrix including life history stage and its interaction with colony
# d_n$samples$tp.lhs <- as.factor(paste(d_n$samples$group, d_n$samples$life_history_stage, sep = "."))
# design_interaction3 <- model.matrix(~ 0+tp.lhs, data = d_n$samples)  # No intercept
# #Clean column names
# colnames(design_interaction3) <- c("Start.Mum", "Start.Pup", "End.Mum", "End.Pup")
# 
# # #voom() converts the read counts to log2-cpm, with associated weights, ready for linear modelling
# vobj_interaction3 <- voom(d_n, design_interaction3, plot=TRUE)
# 
# #Fit a random effect
# dupcor_interaction3 <- duplicateCorrelation(vobj_interaction3, design_interaction3, block = d_n$samples$ID)
# dupcor_interaction3$consensus #the estimated intra individual correlation (0.2734322)
# 
# vobj_interaction3 = voom(d_n, design_interaction3, plot=TRUE, 
#                         block=d_n$samples$ID, correlation=dupcor_interaction3$consensus)
# # #update the correlation for the new voom weights
# dupcor_interaction3 <- duplicateCorrelation(vobj_interaction3, design_interaction3, block = d_n$samples$ID)
# dupcor_interaction3$consensus #the estimated intra individual correlation (0.2733877)
# 
# #Estimate linear mixed model with a single variance component
# vfit_interaction3 <- lmFit(vobj_interaction3, design_interaction3, block=d_n$samples$ID, correlation=dupcor_interaction3$consensus)
# 
# #Fit the contrasts we are interested in
# cont.matrix.interaction3 <- makeContrasts(
#   EndvsStartMum = End.Mum - Start.Mum, #which genes in mothers are DE among tp
#   EndvsStartPup = End.Pup - Start.Pup,  #which genes in pups are DE among tp
#   Interaction = (End.Mum - End.Pup)-(Start.Mum - Start.Pup), #which genes respond differently over time in mums compared to pups
#   levels = colnames(design_interaction3)
# )
# 
# vfit_interaction3 <- contrasts.fit(vfit_interaction3, contrasts=cont.matrix.interaction3)
# 
# #Save the model to avoid recomputing everything
# saveRDS(vfit_interaction3, file = here::here("Models","vfit_interaction3.rds"))

#Load vfit_interaction
vfit_interaction3 <- readRDS(here::here("Models","vfit_interaction3.rds"))
# Fit Empirical Bayes for moderated t-statistics, empirical Bayes moderation is carried out by borrowing information across all the genes to obtain more precise estimates of gene-wise variability
efit_interaction3 <- eBayes(vfit_interaction3)
topTable(efit_interaction3, n=20)
plotSA(efit_interaction3, main="Final model: Mean-variance trend")
plotMD(efit_interaction3)
abline(h=0,col="darkgrey")

#adjusted p-value cutoff that is set at 5% by default.
efit_interaction3$F
#head(efit_interaction3$F.p.value) #pval of the moderated F-statistics (equivalent to one-way ANOVA), it combines all contrast, so in this case it is the same as the t-statistic
#head(efit_interaction3$p.value) #same (not the same?)
summary(decideTests(efit_interaction3, adjust.method="fdr"))

tfit_interaction3 <- treat(vfit_interaction3, lfc=lfc) 
topTreat(tfit_interaction3)
topTreat(tfit_interaction3, coef = "Interaction") #specific for the interaction
nrow(tfit_interaction3)
dt_interaction3 <- decideTests(tfit_interaction3, adjust.method="fdr")
summary(dt_interaction3) 

#Test for significance
#Extract significant genes for each contrast
Allgenes_EndvsStartMum <- as.data.frame(topTreat(tfit_interaction3, coef = "EndvsStartMum", n = nrow(tfit_interaction3)))
Allgenes_EndvsStartPup <- as.data.frame(topTreat(tfit_interaction3, coef = "EndvsStartPup", n = nrow(tfit_interaction3)))
Allgenes_Interaction3 <- as.data.frame(topTreat(tfit_interaction3, coef = "Interaction", n = nrow(tfit_interaction3)))
#Filtered
Filtered_EndvsStartMum <- subset(Allgenes_EndvsStartMum, abs(logFC) >= lfc & adj.P.Val < adj.P.value)
Filtered_EndvsStartPup <- subset(Allgenes_EndvsStartPup, abs(logFC) >= lfc & adj.P.Val < adj.P.value)
Filtered_Interaction3 <- subset(Allgenes_Interaction3, abs(logFC) >= lfc & adj.P.Val < adj.P.value)

Upregulated_EndvsStartMum <- subset(Filtered_EndvsStartMum, logFC > 0)
Downregulated_EndvsStartMum <- subset(Filtered_EndvsStartMum, logFC < 0)

Upregulated_EndvsStartPup <- subset(Filtered_EndvsStartPup, logFC > 0)
Downregulated_EndvsStartPup <- subset(Filtered_EndvsStartPup, logFC < 0)

Upregulated_Interaction3 <- subset(Filtered_Interaction3, logFC > 0)
Downregulated_Interaction3 <- subset(Filtered_Interaction3, logFC < 0)

# Create a list to save the up and down regulated genes and the df containing all genes evaluated for GO analysis
interaction3_list <- list(All_Interaction3 = Allgenes_Interaction3, 
                          Upregulated_interaction3 = Upregulated_Interaction3, 
                          Downregulated_Interaction3 = Downregulated_Interaction3, 
                          Upregulated_EndvsStartPup = Upregulated_EndvsStartPup,
                          Downregulated_EndvsStartPup = Downregulated_EndvsStartPup,
                          Upregulated_EndvsStartMum = Upregulated_EndvsStartMum,
                          Downregulated_EndvsStartMum = Downregulated_EndvsStartMum)

# Loop through and save each data frame using `here()`
for (name in names(interaction3_list)) {
  write.csv(interaction3_list[[name]], file = here::here("GenesGO", paste0(name, ".csv")), row.names = TRUE)
}

#Gene check
#Names <- c("kiaa1522", "fos", "abcc8", "dppa4", "ddit4", "ildr2", "sipa1l2", "arnt2", "crybg2", "gucy2c", "mast4", "pitpnm3", "slc4a8", "tmem150c", "aunip", "chrm3", "mybl1", "kctd19", "ciart")
#Allgenes_EndvsStartMum[Names,]
#Allgenes_EndvsStartPup[Names,]
#Allgenes_Interaction3[Names,]
```
### Heatmap
Creating a heatmap to visualise the differentially expressed genes in the interaction between life history stage and time point. 
```{r}
#| label: load_data
#| echo: FALSE
#Create heatmaps for differentially expressed genes in each contrast
DEG_EndvsStartMum <- rownames(Filtered_EndvsStartMum)
DEG_EndvsStartPup <- rownames(Filtered_EndvsStartPup)
DEG_Interaction3 <- rownames(Filtered_Interaction3)
#Extract the expression values 
expression_EndvsStartMum <- vobj_interaction3$E[DEG_EndvsStartMum, ]
expression_EndvsStartPup <- vobj_interaction3$E[DEG_EndvsStartPup, ]
expression_Interaction3 <- vobj_interaction3$E[DEG_Interaction3, ]

# EndvsStartMum 
#get labels 
annotation_EndvsStartMum <- data.frame(TP = as.factor(d_n$samples$group))
rownames(annotation_EndvsStartMum) <- d_n$samples$Sample
annotation_EndvsStartMum$TP <- ifelse(annotation_EndvsStartMum$TP == 1, "Day 0", "Day 60")
annotation_EndvsStartMum$TP <- factor(annotation_EndvsStartMum$TP, levels = c("Day 0", "Day 60"))

#order the expression values
matrix_EndvsStartMum <- expression_EndvsStartMum[, order(annotation_EndvsStartMum$TP)]
annotation_EndvsStartMum <- annotation_EndvsStartMum[order(annotation_EndvsStartMum$TP), , drop = FALSE]

# EndvsStartPup 
#get labels 
annotation_EndvsStartPup <- data.frame(TP = as.factor(d_n$samples$group))
rownames(annotation_EndvsStartPup) <- d_n$samples$Sample
annotation_EndvsStartPup$TP <- ifelse(annotation_EndvsStartPup$TP == 1, "Day 0", "Day 60")
annotation_EndvsStartPup$TP <- factor(annotation_EndvsStartPup$TP, levels = c("Day 0", "Day 60"))

#order the expression values
matrix_EndvsStartPup <- expression_EndvsStartPup[, order(annotation_EndvsStartPup$TP)]
annotation_EndvsStartPup <- annotation_EndvsStartPup[order(annotation_EndvsStartPup$TP), , drop = FALSE]

# Interaction
annotation_interaction <- data.frame(TP = as.factor(d_n$samples$group),
                                       LHS = as.factor(d_n$samples$life_history_stage))
rownames(annotation_interaction) <- d_n$samples$Sample
annotation_interaction$TP <- ifelse(annotation_interaction$TP == 1, "Day 0", "Day 60")
annotation_interaction$TP <- factor(annotation_interaction$TP, levels = c("Day 0", "Day 60"))
annotation_interaction$LHS <- factor(annotation_interaction$LHS, levels = c("Mother", "Pup"))
annotation_interaction$Group <- interaction(annotation_interaction$LHS, annotation_interaction$TP, sep = "_")
annotation_interaction$Group <- factor(annotation_interaction$Group, levels = c("Mother_Day 0", "Mother_Day 60", "Pup_Day 0", "Pup_Day 60"))

#order the expression values
matrix_interaction <- expression_Interaction3[, order(annotation_interaction$Group)]
annotation_interaction <- annotation_interaction[order(annotation_interaction$Group), , drop = FALSE]
gaps <- cumsum(table(annotation_interaction$Group))
#Drop the group column
annotation_interaction$Group <- NULL 

#Heatmaps
heatmap_EndvsStartMum <- pheatmap(matrix_EndvsStartMum,
                       scale = "row",  # normalize rows (genes)
                       color = colorRampPalette(c("#9D00FF", "white", "#FFA500"))(100),
                       annotation_col = annotation_EndvsStartMum,
                       annotation_colors = ann_colors,
                       cluster_rows = TRUE,
                       treeheight_row = 0,
                       cluster_cols = FALSE,
                       gaps_col = cumsum(table(annotation_EndvsStartMum$TP)),
                       show_rownames = F,
                       show_colnames = F,
                       annotation_legend = FALSE,
                       legend = F,
                       fontsize = 14,
                       fontsize_col = 7,
                       fontsize_row = 6,
                       main = "(a) Interaction: Time Point (Mothers)")

heatmap_EndvsStartPup <- pheatmap(matrix_EndvsStartPup,
                       scale = "row",  # normalize rows (genes)
                       color = colorRampPalette(c("#9D00FF", "white", "#FFA500"))(100),
                       annotation_col = annotation_EndvsStartPup,
                       annotation_colors = ann_colors,
                       cluster_rows = TRUE,
                       treeheight_row = 0,
                       cluster_cols = FALSE,
                       gaps_col = cumsum(table(annotation_EndvsStartPup$TP)),
                       show_rownames = F,
                       show_colnames = F,
                       annotation_legend = FALSE,
                       legend = F,
                       fontsize = 14,
                       fontsize_col = 7,
                       fontsize_row = 6,
                       main = "(b) interaction: Time Point (Pups)")

heatmap_interaction <- pheatmap(matrix_interaction,
                       scale = "row", 
                       color = colorRampPalette(c("#9D00FF", "white", "#FFA500"))(100),
                       annotation_col = annotation_interaction,
                       annotation_colors = ann_colors,
                       cluster_rows = TRUE,
                       treeheight_row = 0,
                       cluster_cols = FALSE,
                       gaps_col = gaps,
                       show_rownames = F,
                       show_colnames = F,
                       annotation_legend = FALSE,
                       legend = F,
                       fontsize = 14,
                       fontsize_col = 7,
                       fontsize_row = 6,
                       main = "(c) Interaction: Time point & life history stage")


ggsave(here::here("Plots","heatmap_EndvsStartMum.png"), plot = heatmap_EndvsStartMum, width = 7, height = 3.25, dpi = 600)
ggsave(here::here("Plots","heatmap_EndvsStartPup.png"), plot = heatmap_EndvsStartPup, width = 7, height = 3.75, dpi = 600)
ggsave(here::here("Plots","heatmap_interaction.png"), plot = heatmap_interaction, width = 7, height = 2, dpi = 600)
```

### Figure of DEG for interaction

```{r}
#| label: load_data
#| echo: FALSE
#Plot DEG

Allgenes_EndvsStartMum <- as.data.frame(topTreat(tfit_interaction3, coef = "EndvsStartMum", n = nrow(tfit_interaction3)))
Allgenes_EndvsStartPup <- as.data.frame(topTreat(tfit_interaction3, coef = "EndvsStartPup", n = nrow(tfit_interaction3)))
Allgenes_Interaction3 <- as.data.frame(topTreat(tfit_interaction3, coef = "Interaction", n = nrow(tfit_interaction3)))


#Significance
Allgenes_EndvsStartMum$Significance <- "'Not significant'"
Allgenes_EndvsStartMum$Significance[Allgenes_EndvsStartMum$adj.P.Val < adj.P.value & Allgenes_EndvsStartMum$logFC >= lfc] <- "Upregulated"
Allgenes_EndvsStartMum$Significance[Allgenes_EndvsStartMum$adj.P.Val < adj.P.value & Allgenes_EndvsStartMum$logFC <= -lfc] <- "Downregulated"

Allgenes_EndvsStartPup$Significance <- "'Not significant'"
Allgenes_EndvsStartPup$Significance[Allgenes_EndvsStartPup$adj.P.Val < adj.P.value & Allgenes_EndvsStartPup$logFC >= lfc] <- "Upregulated"
Allgenes_EndvsStartPup$Significance[Allgenes_EndvsStartPup$adj.P.Val < adj.P.value & Allgenes_EndvsStartPup$logFC <= -lfc] <- "Downregulated"

Allgenes_Interaction3$Significance <- "'Not significant'"
Allgenes_Interaction3$Significance[Allgenes_Interaction3$adj.P.Val < adj.P.value & Allgenes_Interaction3$logFC >= lfc] <- "Upregulated"
Allgenes_Interaction3$Significance[Allgenes_Interaction3$adj.P.Val < adj.P.value & Allgenes_Interaction3$logFC <= -lfc] <- "Downregulated"


ITP_MUM_plot <- ggplot(Allgenes_EndvsStartMum, aes(x = logFC, y = -log10(P.Value), color = Significance, size = Significance)) +
  geom_point() +
  scale_color_manual(values = clrs) +  
  scale_size_manual(values = c(0.5, 1, 1)) +
  labs(title = "(a) Time point (Mothers)", x = "Log2 fold change", y = "-Log10 P-value") +
  theme_classic() +
  ylim(0,30) +
  xlim(-4,4) +
    theme(
    legend.position = "top",
    plot.title.position = "plot")

ITP_Pup_plot <- ggplot(Allgenes_EndvsStartPup, aes(x = logFC, y = -log10(P.Value), color = Significance, size = Significance)) +
  geom_point() +
  scale_color_manual(values = clrs) +  
  scale_size_manual(values = c(0.5, 1, 1)) +
  labs(title = "(b) Time point (Pups)", x = "Log2 fold change", y = "-Log10 P-value") +
  theme_classic() +
  ylim(0,30) +
  xlim(-4,4) +
    theme(
    legend.position = "top",
    plot.title.position = "plot")

Interaction_plot <- ggplot(Allgenes_Interaction3, aes(x = logFC, y = -log10(P.Value), color = Significance, size = Significance)) +
  geom_point() +
  scale_color_manual(values = clrs) +  
  scale_size_manual(values = c(0.5, 1, 1)) +
  labs(title = "(c) Time point & life history stage", x = "Log2 fold change", y = "-Log10 P-value") +
  theme_classic() +
  ylim(0,30) +
  xlim(-4,4) +
    theme(
    legend.position = "top",
    plot.title.position = "plot")

# Combine plots
Interaction_all <- ggarrange(ITP_MUM_plot + fig_theme + theme(axis.title.x = element_blank()), 
                     ITP_Pup_plot + fig_theme + theme(axis.title.x = element_blank(),
                                         axis.title.y = element_blank()), 
                     Interaction_plot + fig_theme, 
                     NULL, common.legend = T, legend = "bottom")

ggsave(here::here("Plots","Interaction_all.png"), plot = Interaction_all, width = 7, height = 7, dpi = 600)
```



## Colony and life stage interaction
Framework for differential gene expression analysis using interactions by ALB. 
```{r}
#| label: load_data
#| echo: FALSE
# #Simple interaction (limma manual 9.5.2)
# #Create a design matrix including life history stage and its interaction with colony
# d_n$samples$colony.lhs <- as.factor(paste(d_n$samples$colony, d_n$samples$life_history_stage, sep = "."))
# design_interaction4 <- model.matrix(~ 0+colony.lhs, data = d_n$samples)  # No intercept
# #Clean column names
# colnames(design_interaction4) <- c("FWB.Mum", "FWB.Pup", "SSB.Mum", "SSB.Pup")
# 
# # #voom() converts the read counts to log2-cpm, with associated weights, ready for linear modelling
# vobj_interaction4 <- voom(d_n, design_interaction4, plot=TRUE)
# 
# #Fit a random effect
# dupcor_interaction4 <- duplicateCorrelation(vobj_interaction4, design_interaction4, block = d_n$samples$ID)
# # #dupcor_interaction4$consensus #the estimated intra individual correlation (0.1308229)
# 
# vobj_interaction4 = voom(d_n, design_interaction4, plot=TRUE, 
#                         block=d_n$samples$ID, correlation=dupcor_interaction4$consensus)
# # #update the correlation for the new voom weights
# dupcor_interaction4 <- duplicateCorrelation(vobj_interaction4, design_interaction4, block = d_n$samples$ID)
# # #dupcor_interaction4$consensus #the estimated intra individual correlation (0.1308384)
# 
# #Estimate linear mixed model with a single variance component
# vfit_interaction4 <- lmFit(vobj_interaction4, design_interaction4, block=d_n$samples$ID, correlation=dupcor_interaction4$consensus)
# 
# #Fit the contrasts we are interested in
# cont.matrix.interaction4 <- makeContrasts(
#   SSBvsFWBMum = SSB.Mum - FWB.Mum, #which genes in mothers are DE among colonies
#   SSBvsFWBPup = SSB.Pup - FWB.Pup,  #which genes in pups are DE among colonies
#   Interaction = (SSB.Mum - SSB.Pup)-(FWB.Mum - FWB.Pup), #which genes respond different in the two colonies in mums compared to pups
#   levels = colnames(design_interaction4)
# )
# 
# vfit_interaction4 <- contrasts.fit(vfit_interaction4, contrasts=cont.matrix.interaction4)
# 
# #Save the model to avoid recomputing everything
# saveRDS(vfit_interaction4, file = here::here("Models","vfit_interaction4.rds"))

#Load vfit_interaction
vfit_interaction4 <- readRDS(here::here("Models","vfit_interaction4.rds"))
# Fit Empirical Bayes for moderated t-statistics, empirical Bayes moderation is carried out by borrowing information across all the genes to obtain more precise estimates of gene-wise variability
efit_interaction4 <- eBayes(vfit_interaction4)
topTable(efit_interaction4, n=20)
plotSA(efit_interaction4, main="Final model: Mean-variance trend")
plotMD(efit_interaction4)
abline(h=0,col="darkgrey")

#adjusted p-value cutoff that is set at 5% by default.
efit_interaction4$F
#head(efit_interaction4$F.p.value) #pval of the moderated F-statistics (equivalent to one-way ANOVA), it combines all contrast, so in this case it is the same as the t-statistic
#head(efit_interaction4$p.value) #same (not the same?)
summary(decideTests(efit_interaction4, adjust.method="fdr"))

tfit_interaction4 <- treat(vfit_interaction4, lfc=lfc) 
topTreat(tfit_interaction4)
topTreat(tfit_interaction4, coef = "Interaction") #specific for the interaction
nrow(tfit_interaction4)
dt_interaction4 <- decideTests(tfit_interaction4, adjust.method="fdr")
summary(dt_interaction4) 
```

## Call in DEG

```{r}
#| label: load_data_for_GO
#| echo: FALSE
#Call up or down regulated genes from differential expression analysis
file_paths <- list.files(path = "GenesGO", pattern = "^(Upregulated|Downregulated).*\\.csv$", full.names = TRUE)

# Loop through and create data frames
for (file in file_paths) {
  name <- tools::file_path_sans_ext(basename(file))      
  df <- read.csv(file) %>%
    mutate(X = toupper(X))
  colnames(df)[1] <- "GeneSymbol"
  assign(name, df, envir = .GlobalEnv)
}

Background_tp <- read.csv("GenesGO/All_tp.csv") %>%
  mutate(X = toupper(X))
colnames(Background_tp)[1] <- "GeneSymbol"

Background_lhs <- read.csv("GenesGO/All_lhs.csv") %>%
  mutate(X = toupper(X))
colnames(Background_lhs)[1] <- "GeneSymbol"

Background_inter <- read.csv("GenesGO/All_Interaction3.csv") %>%
  mutate(X = toupper(X))
colnames(Background_inter)[1] <- "GeneSymbol"

rm(df, file, file_paths, name)
```

#Gene ontology enrichment analysis
```{r}
#| label: Enrichment_analysis
#| echo: FALSE
#Go enrichment analysis for up and down regulated genes
#TP (mothers)
I_TP_MUM_DOWN <- enrichGO(Downregulated_EndvsStartMum$GeneSymbol,
                       OrgDb = org.Hs.eg.db,
                       universe = Background_inter$GeneSymbol,
                       keyType = "SYMBOL",
                       ont = "BP",
                       pvalueCutoff = 0.05,
                       pAdjustMethod = "BH") 
I_TP_MUM_DOWN <- I_TP_MUM_DOWN %>%
  mutate(Regulation = "Downregulated") 

I_TP_MUM_UP <- enrichGO(Upregulated_EndvsStartMum$GeneSymbol,
                       OrgDb = org.Hs.eg.db,
                       universe = Background_inter$GeneSymbol,
                       keyType = "SYMBOL",
                       ont = "BP",
                       pvalueCutoff = 0.05,
                       pAdjustMethod = "BH") 
I_TP_MUM_UP <- I_TP_MUM_UP %>%
  mutate(Regulation = "Upregulated") 

#TP (pups)
I_TP_PUP_DOWN <- enrichGO(Downregulated_EndvsStartPup$GeneSymbol,
                       OrgDb = org.Hs.eg.db,
                       universe = Background_inter$GeneSymbol,
                       keyType = "SYMBOL",
                       ont = "BP",
                       pvalueCutoff = 0.05,
                       pAdjustMethod = "BH") 
I_TP_PUP_DOWN <- I_TP_PUP_DOWN %>%
  mutate(Regulation = "Downregulated") 

I_TP_PUP_UP <- enrichGO(Upregulated_EndvsStartPup$GeneSymbol,
                       OrgDb = org.Hs.eg.db,
                       universe = Background_inter$GeneSymbol,
                       keyType = "SYMBOL",
                       ont = "BP",
                       pvalueCutoff = 0.05,
                       pAdjustMethod = "BH") 
I_TP_PUP_UP <- I_TP_PUP_UP %>%
  mutate(Regulation = "Upregulated") 

#Interaction lhs x tp
I_I_DOWN <- enrichGO(Downregulated_Interaction3$GeneSymbol,
                       OrgDb = org.Hs.eg.db,
                       universe = Background_inter$GeneSymbol,
                       keyType = "SYMBOL",
                       ont = "BP",
                       pvalueCutoff = 0.05,
                       pAdjustMethod = "BH") 
I_I_DOWN <- I_I_DOWN %>%
  mutate(Regulation = "Downregulated") 

I_I_UP <- enrichGO(Upregulated_interaction3$GeneSymbol,
                       OrgDb = org.Hs.eg.db,
                       universe = Background_inter$GeneSymbol,
                       keyType = "SYMBOL",
                       ont = "BP",
                       pvalueCutoff = 0.05,
                       pAdjustMethod = "BH") 
I_I_UP <- I_I_UP %>%
  mutate(Regulation = "Upregulated") 
```

## Top 20 enriched GOterms (biological process) 

### Interaction
```{r}
#| label: load_data
#| echo: FALSE
#Combine
I_TP_MUM <- rbind(as.data.frame(I_TP_MUM_DOWN), as.data.frame(I_TP_MUM_UP)) %>%
  arrange(p.adjust) 

I_TP_PUP <- rbind(as.data.frame(I_TP_PUP_DOWN), as.data.frame(I_TP_PUP_UP)) %>%
  arrange(p.adjust) 

I_I <- rbind(as.data.frame(I_I_DOWN), as.data.frame(I_I_UP)) %>%
  arrange(p.adjust) 

#Identify top20
GO_I_TP_MUM_Top20 <- I_TP_MUM %>%
  arrange(p.adjust) %>%
  slice_head(n = 20) 

GO_I_TP_PUPTop20 <- I_TP_PUP %>%
  arrange(p.adjust) %>%
  slice_head(n = 20) 

GO_I_I_Top20 <- I_I %>%
  arrange(p.adjust) %>%
  slice_head(n = 20) 

#Top 20 GO terms and description
GO_I_TP_MUM_Top20_table <- GO_I_TP_MUM_Top20[, c("ID", "Description")]
colnames(GO_I_TP_MUM_Top20_table) <- c("GO Term", "Description")
rownames(GO_I_TP_MUM_Top20_table) <- NULL
#Force line break
GO_I_TP_MUM_Top20_table$Description <- str_wrap(GO_I_TP_MUM_Top20_table$Description, width = 45)

GO_I_TP_PUP_Top20_table <- GO_I_TP_PUPTop20[, c("ID", "Description")]
colnames(GO_I_TP_PUP_Top20_table) <- c("GO Term", "Description")
rownames(GO_I_TP_PUP_Top20_table) <- NULL
#Force line break
GO_I_TP_PUP_Top20_table$Description <- str_wrap(GO_I_TP_PUP_Top20_table$Description, width = 45)

GO_I_I_Top20_table <- GO_I_I_Top20[, c("ID", "Description")]
colnames(GO_I_I_Top20_table) <- c("GO Term", "Description")
rownames(GO_I_I_Top20_table) <- NULL
#Force line break
GO_I_I_Top20_table$Description <- str_wrap(GO_I_I_Top20_table$Description, width = 45)


#Fancy smancy bubble plot
plot_I_TP_MUM <- ggplot(I_TP_MUM, aes(x = zScore, y = -log10(p.adjust))) +
  geom_point(aes(size = Count, shape = Regulation, fill = Regulation), color = "black", alpha = 0.7) +
  scale_shape_manual(values = c("Upregulated" = 24, "Downregulated" = 25)) +  
  scale_fill_manual(values = c("Upregulated" = "#FDA172", "Downregulated" = "#9867B5")) +
  scale_size_continuous(name = "DEG Count") +
  #Add the top terms to the plot
  geom_text_repel(
  data = GO_I_TP_MUM_Top20,  
  aes(label = ID),
  nudge_y = -0.03,     
  size = 3,
  segment.color = NA,
  #force = 2,
  max.overlaps = Inf
) +
  labs(
    title = "Interaction: Time point (Mothers)",
    x = "Z-score",
    y = expression(-log[10]("Adjusted p-value (FDR)")),
    shape = "Regulation"
  ) +
  theme_classic() +
  theme(legend.position = "right",
        legend.title = element_text(size = 14),   
        legend.text = element_text(size = 12),     
        legend.key.size = unit(0.4, "cm")) +
  guides(
  size = guide_legend(
    override.aes = list(shape = 24)  # use triangle in size legend
  ),
  fill = "none")

plot_I_TP_PUP <- ggplot(I_TP_PUP, aes(x = zScore, y = -log10(p.adjust))) +
  geom_point(aes(size = Count, shape = Regulation, fill = Regulation), color = "black", alpha = 0.7) +
  scale_shape_manual(values = c("Upregulated" = 24, "Downregulated" = 25)) +  
  scale_fill_manual(values = c("Upregulated" = "#FDA172", "Downregulated" = "#9867B5")) +
  scale_size_continuous(name = "DEG Count") +
  #Add the top terms to the plot
  geom_text_repel(
  data = GO_I_TP_PUPTop20,  
  aes(label = ID),
  nudge_y = -0.03,     
  size = 3,
  segment.color = NA,
  #force = 2,
  max.overlaps = Inf
) +
  labs(
    title = "Interaction: Time point (Pups)",
    x = "Z-score",
    y = expression(-log[10]("Adjusted p-value (FDR)")),
    shape = "Regulation"
  ) +
  theme_classic() +
  theme(legend.position = "right",
        legend.title = element_text(size = 14),   
        legend.text = element_text(size = 12),     
        legend.key.size = unit(0.4, "cm")) +
  guides(
  size = guide_legend(
    override.aes = list(shape = 24)  # use triangle in size legend
  ),
  fill = "none")

plot_I_I <- ggplot(I_I, aes(x = zScore, y = -log10(p.adjust))) +
  geom_point(aes(size = Count, shape = Regulation, fill = Regulation), color = "black", alpha = 0.7) +
  scale_shape_manual(values = c("Upregulated" = 24, "Downregulated" = 25)) +  
  scale_fill_manual(values = c("Upregulated" = "#FDA172", "Downregulated" = "#9867B5")) +
  scale_size_continuous(name = "DEG Count") +
  #Add the top terms to the plot
  geom_text_repel(
  data = GO_I_I_Top20,  
  aes(label = ID),
  nudge_y = -0.03,     
  size = 3,
  segment.color = NA,
  force = 2,
  max.overlaps = Inf
) +
  labs(
    title = "Interaction: Time point & life history stage)",
    x = "Z-score",
    y = expression(-log[10]("Adjusted p-value (FDR)")),
    shape = "Regulation"
  ) +
  theme_classic() +
  theme(legend.position = "right",
        legend.title = element_text(size = 14),   
        legend.text = element_text(size = 12),     
        legend.key.size = unit(0.4, "cm")) +
  guides(
  size = guide_legend(
    override.aes = list(shape = 24)  # use triangle in size legend
  ),
  fill = "none")

## Tables
# Create alternating row colors
row_colors <- rep(c("white", "#f0f0f0"), length.out = nrow(GO_LHS_Top20_table))

p_table_I_TP_MUM <- ggtexttable(
  GO_I_TP_MUM_Top20_table,
  rows = NULL,
  theme = ttheme(
    "light",
    tbody = list(bg_params = list(fill = row_colors))))

p_table_I_TP_PUP <- ggtexttable(
  GO_I_TP_PUP_Top20_table,
  rows = NULL,
  theme = ttheme(
    "light",
    tbody = list(bg_params = list(fill = row_colors))))

p_table_I_I <- ggtexttable(
  GO_I_I_Top20_table,
  rows = NULL,
  theme = ttheme(
    "light",
    tbody = list(bg_params = list(fill = row_colors))))

EA_I_TP_MUM <- plot_grid(
 plot_I_TP_MUM,
 NULL,
 p_table_I_TP_MUM,
 ncol = 3,
 rel_widths = c(1.3,0.01,0.9)
)

EA_I_TP_PUP <- plot_grid(
 plot_I_TP_PUP,
 NULL,
 p_table_I_TP_PUP,
 ncol = 3,
 rel_widths = c(1.3,0.01,0.9)
)

EA_I_I <- plot_grid(
 plot_I_I,
 NULL,
 p_table_I_I,
 ncol = 3,
 rel_widths = c(1.3,0.01,0.9)
)

ggsave(here::here("Plots","EA_I_TP_MUM.png"), plot = EA_I_TP_MUM, width = 12, height = 7.5, dpi = 600)
ggsave(here::here("Plots","EA_I_TP_PUP.png"), plot = EA_I_TP_PUP, width = 12, height = 7.5, dpi = 600)
ggsave(here::here("Plots","EA_I_I.png"), plot = EA_I_I, width = 12, height = 7.5, dpi = 600)
```

## Revigo in R
REduce and VIsualise Gene Ontology - a tool to reduce redundant GO terms and group the terms based on semantic similarity. 
```{r}
#| label: load_data
#| echo: FALSE
remotes::install_github("ssayols/rrvgo")
library(rrvgo)
```

### Similarity matrices

```{r}
#| label: Similarity_matrices
#| echo: FALSE
#calculate similarity matrices for all contrasts
SM_I_TP_MUM_DOWN <- calculateSimMatrix(I_TP_MUM_DOWN$ID, orgdb = "org.Hs.eg.db", ont = "BP", method = "Rel")
SM_I_TP_MUM_UP <- calculateSimMatrix(I_TP_MUM_UP$ID, orgdb = "org.Hs.eg.db", ont = "BP", method = "Rel")
SM_I_TP_PUP_DOWN <- calculateSimMatrix(I_TP_PUP_DOWN$ID, orgdb = "org.Hs.eg.db", ont = "BP", method = "Rel")
SM_I_TP_PUP_UP <- calculateSimMatrix(I_TP_PUP_UP$ID, orgdb = "org.Hs.eg.db", ont = "BP", method = "Rel")
SM_I_I_DOWN <- calculateSimMatrix(I_I_DOWN$ID, orgdb = "org.Hs.eg.db", ont = "BP", method = "Rel")
SM_I_I_UP <- calculateSimMatrix(I_I_UP$ID, orgdb = "org.Hs.eg.db", ont = "BP", method = "Rel")
```

### Reduce terms
```{r}
#| label: Reduce_terms
#| echo: FALSE

## Interaction
scores_I_TP_MUM_DOWN <- -log10(as.numeric(I_TP_MUM_DOWN$p.adjust))
names(scores_I_TP_MUM_DOWN) <- I_TP_MUM_DOWN$ID
scores_I_TP_MUM_UP <- -log10(as.numeric(I_TP_MUM_UP$p.adjust))
names(scores_I_TP_MUM_UP) <- I_TP_MUM_UP$ID
scores_I_TP_PUP_DOWN <- -log10(as.numeric(I_TP_PUP_DOWN$p.adjust))
names(scores_I_TP_PUP_DOWN) <- I_TP_PUP_DOWN$ID
scores_I_TP_PUP_UP <- -log10(as.numeric(I_TP_PUP_UP$p.adjust))
names(scores_I_TP_PUP_UP) <- I_TP_PUP_UP$ID
scores_I_I_DOWN <- -log10(as.numeric(I_I_DOWN$p.adjust))
names(scores_I_I_DOWN) <- I_I_DOWN$ID
scores_I_I_UP <- -log10(as.numeric(I_I_UP$p.adjust))
names(scores_I_I_UP) <- I_I_UP$ID

#Reduce terms in interaction
reduced_I_TP_MUM_DOWN <- reduceSimMatrix(
    simMatrix = SM_I_TP_MUM_DOWN,
    scores = scores_I_TP_MUM_DOWN,
    threshold = 0.7,
    orgdb = "org.Hs.eg.db")

reduced_I_TP_MUM_UP <- reduceSimMatrix(
    simMatrix = SM_I_TP_MUM_UP,
    scores = scores_I_TP_MUM_UP,
    threshold = 0.7,
    orgdb = "org.Hs.eg.db")

reduced_I_TP_PUP_DOWN <- reduceSimMatrix(
    simMatrix = SM_I_TP_PUP_DOWN,
    scores = scores_I_TP_PUP_DOWN,
    threshold = 0.7,
    orgdb = "org.Hs.eg.db")

reduced_I_TP_PUP_UP <- reduceSimMatrix(
    simMatrix = SM_I_TP_PUP_UP,
    scores = scores_I_TP_PUP_UP,
    threshold = 0.7,
    orgdb = "org.Hs.eg.db")

#Tagged out because it cannot reduce the terms further. 
#reduced_I_I_DOWN <- reduceSimMatrix(
#    simMatrix = SM_I_I_DOWN,
#    scores = scores_I_I_DOWN,
#    threshold = 0.7,
#    orgdb = "org.Hs.eg.db")

reduced_I_I_UP <- reduceSimMatrix(
    simMatrix = SM_I_I_UP,
    scores = scores_I_I_UP,
    threshold = 0.7,
    orgdb = "org.Hs.eg.db")
```

## Heatmaps
### Interaction
```{r}
#| label: revigo_heatmap_interaction
#| echo: FALSE

HP_I_TP_MUM_DOWN <- heatmapPlot(SM_I_TP_MUM_DOWN,
                                  reduced_I_TP_MUM_DOWN,
                                  annotateParent=TRUE,
                                  annotationLabel="parentTerm",
                                  fontsize=6,
                                 main = "Interaction: Time point - Mothers (Downregulated)")

HP_I_TP_MUM_UP <- heatmapPlot(SM_I_TP_MUM_UP,
                                  reduced_I_TP_MUM_UP,
                                  annotateParent=TRUE,
                                  annotationLabel="parentTerm",
                                  fontsize=6,
                                  main = "Interaction: Time point - Mothers (Upregulated)")

HP_GO_I_TP_MUM <- plot_grid(HP_I_TP_MUM_UP$gtable, HP_I_TP_MUM_DOWN$gtable, ncol = 1)
ggsave(here::here("Plots","HP_GO_I_TP_MUM.png"), plot = HP_GO_I_TP_MUM, width = 7, height = 10, dpi = 600)

HP_I_TP_PUP_DOWN <- heatmapPlot(SM_I_TP_PUP_DOWN,
                                  reduced_I_TP_PUP_DOWN,
                                  annotateParent=TRUE,
                                  annotationLabel="parentTerm",
                                  fontsize=6,
                                 main = "Interaction: Time point - Pups (Downregulated)")

HP_I_TP_PUP_UP <- heatmapPlot(SM_I_TP_PUP_UP,
                                  reduced_I_TP_PUP_UP,
                                  annotateParent=TRUE,
                                  annotationLabel="parentTerm",
                                  fontsize=6,
                                  main = "Interaction: Time point - Pups (Upregulated)")

HP_GO_I_TP_PUP <- plot_grid(HP_I_TP_PUP_UP$gtable, HP_I_TP_PUP_DOWN$gtable, ncol = 1)
ggsave(here::here("Plots","HP_GO_I_TP_PUP.png"), plot = HP_GO_I_TP_PUP, width = 7, height = 10, dpi = 600)

HP_I_I_UP <- heatmapPlot(SM_I_I_UP,
                                  reduced_I_I_UP,
                                  annotateParent=TRUE,
                                  annotationLabel="parentTerm",
                                  fontsize=6,
                                  main = "Interaction: Time point & life history stage (Upregulated)")

HP_I_I_UP_plot <- plot_grid(HP_I_I_UP$gtable)
ggsave(here::here("Plots","HP_I_I_UP_plot.png"), plot = HP_I_I_UP_plot, width = 9, height = 5, dpi = 600)
```

## Scatterplots
### Interaction
```{r}
#| label: scatterplot_interaction
#| echo: FALSE


SP_I_TP_MUM_DOWN <- scatterPlot(SM_I_TP_MUM_DOWN,reduced_I_TP_MUM_DOWN) +  
                                theme_classic() + 
                                xlim(c(-1,1)) + 
                                ylim(c(-1,1)) +
                                xlab("Semantic space x") +
                                ylab("Semantic space y") +
                                ggtitle("Interaction: Time point - Mothers (Downregulated)")

SP_I_TP_MUM_UP <- scatterPlot(SM_I_TP_MUM_UP,reduced_I_TP_MUM_UP) +  
                                theme_classic() + 
                                xlim(c(-1,1)) + 
                                ylim(c(-1,1)) +
                                xlab("Semantic space x") +
                                ylab("Semantic space y") +
                                ggtitle("Interaction: Time point - Mothers (Upregulated)")

SP_GO_I_TP_MUM <- plot_grid(SP_I_TP_MUM_UP, SP_I_TP_MUM_DOWN)
ggsave(here::here("Plots","SP_GO_I_TP_MUM.png"), plot = SP_GO_I_TP_MUM, width = 9, height = 4.5, dpi = 600)

SP_I_TP_PUP_DOWN <- scatterPlot(SM_I_TP_PUP_DOWN,reduced_I_TP_PUP_DOWN) +  
                                theme_classic() + 
                                xlim(c(-1,1)) + 
                                ylim(c(-1,1)) +
                                xlab("Semantic space x") +
                                ylab("Semantic space y") +
                                ggtitle("Interaction: Time point - Pups (Downregulated)")

SP_I_TP_PUP_UP <- scatterPlot(SM_I_TP_PUP_UP,reduced_I_TP_PUP_UP) +  
                                theme_classic() + 
                                xlim(c(-1,1)) + 
                                ylim(c(-1,1)) +
                                xlab("Semantic space x") +
                                ylab("Semantic space y") +
                                ggtitle("Interaction: Time point - Pups (Upregulated)")

SP_GO_I_TP_PUP <- plot_grid(SP_I_TP_PUP_UP, SP_I_TP_PUP_DOWN)
ggsave(here::here("Plots","SP_GO_I_TP_PUP.png"), plot = SP_GO_I_TP_PUP, width = 9, height = 4.5, dpi = 600)


SP_I_I_UP <- scatterPlot(SM_I_I_UP,reduced_I_I_UP) +  
                                theme_classic() + 
                                xlim(c(-1,1)) + 
                                ylim(c(-1,1)) +
                                xlab("Semantic space x") +
                                ylab("Semantic space y") +
                                ggtitle("Interaction: Time point & life history stage (Upregulated)")

SP_I_I_UP_plot <- plot_grid(SP_I_I_UP)
ggsave(here::here("Plots","SP_I_I_UP_plot.png"), plot = SP_I_I_UP_plot, width = 4.5, height = 4.5, dpi = 600)
```